### 2023-11-28 response/rebuttal.tex
create rebuttal.tex

% LaTeX rebuttal letter example. 
% 
% Copyright 2019 Friedemann Zenke, zenkelab.org
%
% Based on examples by Dirk Eddelbuettel, Fran and others from 
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
% 
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{P\,\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent 
	{\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newenvironment{changes}
   {\medskip \noindent \begin{sf}\textbf{Changes\_made}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
	\ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
	\medskip \end{sf}}

\begin{document}

\section*{Response to the reviewers, TOSEM-2023-0120}
% General intro text goes here
We thank the reviewers for their critical assessment of our work. 
In the following we address their concerns point by point.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description 
\begin{point}
Figure 1 shows a sequence of transplatations, which I agree is easy to represent for such kind of succession, but is the order really important? In my opinion, probably not, except maybe when analysing the underlying dependency graph? Could it be interesting to analyse the impact of this analysis on missed artefacts when using ProdScalpel? (maybe to relate to reusability in compilation)
	\label{pt:foo}
\end{point}
\begin{reply}
TODO
\end{reply}

\begin{point}
Section 3.2, Page 9, "To overcome these issues Foundry introduces an organ-host wrapper. This layer is responsible for providing access to the organ from the target host. It is automatically constructed on demand, according to a given implantation point in the product base."+
when I saw this, I wonder whether Foundry would only produce wrappers to call new features and thus, would it be possible that the product line becomes polygot (in the sense that the wrapper may only be a call to "exec" for instance whatever the program called)? 
It is probably not as simple as that, but could it be a viable solution? What would be the impacts on the host system (in term of performance for instance)?
	\label{pt:foo}
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point}
Section 4.1, page 12, "Figure 8 gives a example of a portion of code after prodScalpel cleaned up unused directives."
in the end, from what I can see from Figure 8, ProdScalpel is able to provide two output "partial source code", one to use when the feature is activated and the second one when it is not, when adding a new feature to the host. The choice between the two will depend on the new product to derive I assume. Is that correct?
	\label{pt:foo}
\end{point}
\begin{reply}
	TODO
\end{reply}
\begin{changes}
    TODO
\end{changes}

\begin{point}
Section 5.2, page 15, when authors use VI and VIM both as donor and product base. Did authors took two different versions of VI (and VIM)? One with the desired feature and one without it. That could make sense to the why trying to transplant a feature to the same codebase. It could also be a maner to evaluate the tool. Is it able to provide the same code in the end than the original? How does it differ? Which tests pass or fail? etc.
	\label{pt:foo}
\end{point}
\begin{reply}
	TODO
\end{reply}
\begin{changes}
    TODO
\end{changes}

\begin{point}
Section 5.3, page 16, when answering RQ2, I have the feeling that it lacks some in-depth analysis. From my understanding, I assume that some products cannot be run due to missed instructions or something else, but it is not discussed. The text only refers to documentation that is imprecise.
In addition, maybe RQ2 is expressed in a misleading way when put in light of the analysis. I was expected to get a taxonomy of different features that can or cannot be handled yet.

	\label{pt:foo}
\end{point}

\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\subsection*{Minor}

\shortpoint{Section 2, Page 4, line 20 "a document viewer for multiple document formats that is part of GNOME, not not an editor." -> an additional "not" maybe it was a "but" in the first place.}
\shortreply{ TODO.}

\shortpoint{Figure 3 mentions two products P and Q but no products are visible on the figure.}
\shortreply{ TODO.}

\shortpoint{I wonder whether Figures 1, 2, and 3 somehow may be redundant. Could they be merged? Is it a necessity to keep all these 3? (at least 1 and 2 separated?)}
\shortreply{ TODO.}

% Begin a new reviewer section
\reviewersection

\begin{point}

1) The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
\end{point}
\begin{reply}
	We appreciate the reviewer's thoughtful consideration of our paper. We acknowledge the concern raised regarding the clarity of the problem definition and its alignment with the provided example. We understand the importance of a well-defined problem statement for a thorough evaluation of our proposed solution.

The central problem we address in our paper revolves around the challenges inherent in Software Product Line (SPL) re-engineering. Specifically, we focus on the need for a systematic and automated approach to transplant reusable components (organs) from existing SPLs into a new product base. This process, which we term "software transplantation," aims to streamline SPL re-engineering by enhancing reusability and reducing redundancy.

To provide a concrete illustration, we offer an example in the paper to elucidate the transplantation process. However, we acknowledge that there might be room for improvement in aligning this example more closely with the general problem definition. We revisited the example in our manuscript to ensure better coherence with the overarching problem statement.

In Section 3, where we detail our software transplantation approach, Foundry, we will provide additional context to clarify the problem definition further. This clarification include a refined articulation of the challenges associated with SPL re-engineering that our approach aims to address. By doing so, we wait had enhanced the overall coherence between the problem definition and the proposed solution.

We appreciate the reviewer's guidance on this matter, and we are committed to enhancing the clarity of our problem statement to facilitate a more accurate evaluation of our solution.

Thank you for your valuable feedback, and we look forward to addressing any further concerns you may have.
\end{reply}
\begin{changes}
    TODO
\end{changes}

\begin{point} 
2) The solution is described in a vague way. A more theoretical or at least rigorous approach should be followed to describe the solution. The solution seems to be more like a white paper for a tool than a research paper with reproducible elements. 
\end{point}
\begin{reply}
TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
1) In the abstract, you say that this proposal is to re-engineer existing product into a SPL. However, the motivating example and the solution does not fit well with this. In Section 2 you say that the GNOME team could be interested in offering an SPL of text editors. Then, you describe two other tools where you could take features from and "transplant"  it to a common code base generated from Gedit. This is confusing because when you are in an extractive SPL approach you have a set of products and you do not know from which one you should start building the code base for your SPL. Why Gedit and not Evince or Nedit? Are really those systems of a family of systems managed from the same organization or are they just individual products that share some common features. 
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
The example seems to be more on "copying" features from other products rather than re-engineering an existing SPL. Also, you present in Fig. 1 the donors and the  host using a feature model. Are they already SPLs according to that? Are you re-engineering an SPL from existing SPLs? All this is confusing and makes it difficult to understand the problem you want to solve. The problem has to be very well explained so that the solution can be evaluated. 
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
Also, it is confusing in the example that you talk about three editors (gedit, evince and nedit). However, in the introduction you say that for evaluating your proposal you used other text editors (kilo, vi and cflow on top of vi and vim). This is confusing and distracts the reader. 
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
2) The solution is weakly presented. The difference between "foundary" and "prodscalpel" is confusing. After some reads, one can understand that one if the conceptual and the other is the implementation. However, this should be cleared right from the beginning. Nevertheless, the main problem comes when the solution is described. Section 3 is not formally or rigorously described. There is no a single definition along the sections and only descriptive paragraphs are used. This is unusual in prestigious journals such as TOSEM. The way you describe your proposal is more for a workshop to present ideas rather to a solid result to be presented in a journal. 
\end{point}

\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
Why in Figure 2 system 4 is selected? why P and Q are the "code base" and feature N and O are removed? P and Q are not features that are shared with other systems, are they?
\end{point}

\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
In addition, the implementation is described as a tool paper and adds no value to rigorously and clearly present your approach/tool. 
\end{point}

\begin{reply}
	TODO
\end{reply}

\begin{changes}
    TODO
\end{changes}

\begin{point} 
- The title is confusing. It seems that the paper is more about "feature extraction and implantation using software transplantation" and it is not much about SPLE
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
In the abstract, the ideas are not clear. More details should be given. What is your solution about? In what is based? Only high level descriptions are given.
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
Add a reference to support the first sentence of the 2nd paragraph of Section 1
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
The first time that prodscalpel is mentioned is in line 44 (pag 3 of 32) and it is confusing that it is. 
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
Fig 1 is in page 4 but the first mention to it is in pag 5 and in another section. This is confusing. 
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
Foundary is described as "an approach" (l 29 p 4). It would be nice to more precisely describe it: process? methodology fragment? tool? what?
\end{point}
\begin{reply}
	TODO
\end{reply}

\begin{point} 
The statement that GNOME is a natural candidate for SPL is quite strong! (l 9, p5)
\end{point}

\end{document}

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/abbreviations.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

% English
\newcommand{\aka}{\hbox{\emph{aka}}\xspace}
\newcommand{\cf}{\hbox{\emph{cf.}}\xspace}
\newcommand{\deletia}{\ldots [deletia] \ldots}
\newcommand{\etal}{\hbox{\emph{et al.}}\xspace}
\newcommand{\eg}{\hbox{\emph{e.g.}}\xspace}
\newcommand{\etc}{\hbox{\emph{etc.}}\xspace}
\newcommand{\ie}{\hbox{\emph{i.e.}}\xspace}
\newcommand{\scil}{\hbox{\emph{sc.}}\xspace} %scilicet: it is permitted to know
%\newcommand{\st}{\hbox{\emph{s.t.}}\xspace}
\newcommand{\viz}{\hbox{\emph{viz.}}\xspace} %videlicet: it is permitted to see
\newcommand{\wrt}{\hbox{\emph{w.r.t.}}\xspace}

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/main.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

%\documentclass[manuscript,screen]{acmart}
\documentclass[acmtacm,screen]{acmart}
%\documentclass[acmsmall, screen, review=false]{acmart}
%\documentclass[anonymous,print]{acmart}
%\usepackage[true]{anonymous-acm}
\usepackage[nameinlink]{cleveref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{framed}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{adjustbox}
\input{abbreviations.tex}

% hyperref commands to correct capitalise \autoref labels
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}
\def\algorithmautorefname{Algorithm}
%github
\usepackage[switch]{lineno}%line number
%\linenumbers
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%\setcopyright{acmcopyright}
%\copyrightyear{2022}
%\acmYear{2022}
%\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a JOURNAL article.
%\acmJournal{JACM}
%\acmVolume{37}
%\acmNumber{4}
%\acmArticle{111}
%\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Comments %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{ifthen}
\usepackage{xcolor}

\newboolean{showcomments}
\setboolean{showcomments}{true} % comment this line to deactivate comments
\ifthenelse{\boolean{showcomments}}{
  \newcommand{\nbc}[3]{
    {\colorbox{#3}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}%
    {\textcolor{#3}{\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}}}
  \newcommand{\todo}[1]{\nbc{TODO}{#1}{blue}\xspace}
}{
  \newcommand{\nbc}[3]{}
  \newcommand{\todo}[1]{}
}
\newcommand{\eb}[1]{\nbc{Earl}{#1}{orange}}
\newcommand{\jp}[1]{\nbc{Justyna}{#1}{blue}}
\newcommand{\ls}[1]{\nbc{Leandro}{#1}{red}}
\newcommand\re[1]{\nbc{To Review:}{#1}{green}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% End Comments %%%%%%%%%%%%%%%%%%%%%%%%%

%%% Project Macros
\newcommand{\FOUNDRY}{\textsc{Foundry}\xspace}
\newcommand{\prodscalpel}{\textsc{prodScalpel}\xspace}
\newcommand{\projURL}{\url{utopia.com}}

\begin{document}

\title{Software Product Line Engineering via Software Transplantation }

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Leandro O. Souza}
\email{leandro.souza@ifba.edu.br}
\orcid{0009-0001-6264-2573}
%\orcid{1234-5678-9012}
%\authornotemark[1]
\affiliation{%
  \institution{Federal Institute of Bahia}
 % \streetaddress{Rua Rogério Nunes Dourado, 27}
  \city{Irecê}
  \state{Bahia}
  \country{Brazil}
 % \postcode{44900-000}
}

\author{Earl T. Barr}
\email{E.Barr@ucl.ac.uk}
\orcid{0000-0003-0771-7891}
\affiliation{
    \institution{University College London}
    \city{London}
    \country{UK}
    }

\author{Justyna Petke}
\email{j.petke@ucl.ac.uk}
\orcid{0000-0002-7833-6044}
\affiliation{
    \institution{University College London}
    \city{London}
    \country{UK}
    }

\author{Eduardo S. Almeida}
\email{esa@rise.com.br}
\affiliation{
    \institution{Federal University of Bahia}
    \city{Salvador}
    \state{Bahia}
    \country{Brazil}
    }

\author{Paulo Anselmo M. S. Neto}
\email{paulo.motant@ufrpe.br}
\orcid{0000-0003-0197-8249}
\affiliation{
    \institution{Federal Rural University of Pernambuco}
    \city{Recife}
    \state{Pernambuco}
    \country{Brazil}
    }

%\author{ADD AUTHOR NAMES}
%\email{leandro.souza@ifba.edu.br}
%\affiliation{%
%  \institution{University College London}
  %\streetaddress{1 Th{\o}rv{\"a}ld Circle}
%  \city{London}
%  \country{UK}}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

\begin{abstract}
For companies producing related products, a Software Product Line (SPL) is a software reuse method that improves time-to-market and software quality, achieving substantial cost reductions. 
These benefits do not come for free.
It often takes years to re-architect and re-engineer a codebase to support SPL and, once adopted, it must be maintained. Current SPL practice relies on a collection of tools, tailored for different reengineering phases, whose output developers must coordinate and integrate. 
We present \FOUNDRY, a general automated approach for leveraging software transplantation to speed conversion to and maintenance of SPL. 
\FOUNDRY facilitates feature extraction and migration. 
It can efficiently, repeatedly, transplant a sequence of features, implemented in multiple files. 
We used \FOUNDRY to create two valid product lines that integrate features from three real-world systems in an automated way. 
Moreover, we conducted an experiment comparing \FOUNDRY's feature migration with manual effort. We show that \FOUNDRY automatically migrated features across codebases 4.8 times faster, on average, than the average time a group of SPL experts took to accomplish the task.
\end{abstract}

%We show that \FOUNDRY takes, on average, 18\% of the time to reuse features across codebases than a group of SPL experts.
%\end{abstract}
%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
% ADD CCS CONCEPTS

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Software Product Lines, Software Transplantation, Genetic Improvement}

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
%sections
\input{sections/sec01-introduction}
\label{intro}

\input{sections/sec02-motivating_example.tex}
\label{sec:motivating_example}

\input{sections/sec03-approach.tex}
\label{sec:approach}

\input{sections/sec04-implementation.tex}
\label{sec:implementation}

\input{sections/sec06-case_studies.tex}
\label{sec:case_studies}

\input{sections/sec07-experiment.tex}
\label{sec:experiment}

\input{sections/sec05-automated_spl_reengineering.tex}
\label{sec:automated_spl_reengineering}

\input{sections/sec08-related_work.tex}
\label{sec:related_work}

\input{sections/sec09-conclusion.tex}
\label{sec:conclusion} 
%-------------

%\section{Acknowledgments}

%This work was support by the UKRI EPSRC grant no. EP/P023991/1.

\section{Copyright}

For the purpose of open access, the author has applied a Creative Commons Attribution (CC BY) licence to any Author Accepted Manuscript version arising.

%\section{Appendices}

%If your work needs an appendix, add it before the ``\verb|\end{document}|'' command at the conclusion of your source document.

%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio.bib}

%% If your work has an appendix, this is the place to put it.
\appendix

\end{document}
\endinput

%% End of file `main.tex'.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec01-introduction.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Introduction}
\label{sec:introduction}

Software Product Line (SPL) is a systematic methodology for producing related software products from shared development assets~\cite{Pohl2005, Linden2007, Fischer2015}.
%SPL organises a codebase into a core asset features that are shared across all products within the product line. Then, it defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
SPL organises features as core assets that are shared across all products within the product line. Then, it defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
%SPL organises a codebase into a \emph{product base}, a set of  features, then defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
By centralising features into a product base, SPL prevents cross-product feature drift, allowing all products to benefit from a feature's improvement, if they use it.
For companies producing related products, adopting SPL improves productivity and quality, speeds time to market, and reduces cost, because it facilitates the reuse of development artifacts, such as code and design~\cite{Linden2007,Bastos2015}.

Despite its benefits, adopting SPL requires considerable upfront investment before its benefits can be realised. 
The cost of migrating existing products to SPL is lower than adopting SPL from scratch, making \emph{extractive}~\cite{Krueger2001} adoption more common, especially in companies with many software system variants in production~\cite{Berger2013}. Two factors drive this preference: 1) it is often hard to know upfront that SPL will be needed because related products often emerge from a small set of initial products, and 2)  starting from scratch discards considerable knowledge and investment in existing codebases, when they exist~\cite{Breivold2008,Northrop2012}.

To re-engineer existing products, companies must solve four problems:  They must analyse their products to 1) identify and 2) extract the features these products share, and 3) learn their inter-dependencies. Finally, they must 4) define a variability mechanism for combining these features, subject to their inter-dependency constraints~\cite{Assuncao2017}. 
Currently, reengineering to adopt SPL remains largely manual~\cite{Assuncao2017} and costly~\cite{Bockle2004}. Indeed, because of its cost, software companies delay, or even refrain from, adopting SPL~\cite{Fischer2015}.  Automating these tasks remains an open challenge~\cite{Assuncao2017}.

%\textcolor{magenta}{From Justyna to Leandro: I am missing here a paragraph that would summarise related work on automating reengineering tasks.}\ls{fixed}

In 2013, Harman et al.~\cite{Harman2013} introduced software transplantation (ST) as a new research direction and laid out its implications for SPL reengineering. Harman et al. defined software transplantation as ``\emph{the adaptation of one system's behaviour or structure to incorporate a subset of the behaviour or structure of another}''~\cite{Harman2013}. In terms of automated software transplantation, Petke et al.~\cite{Petke2014,Petke2018} were the pioneers in transplanting code snippets from different versions of a system to enhance its performance using genetic improvement~\cite{Petke18}. A year later,  Barr et al.~\cite{Barr2015} introduced a theory, algorithm, and tool that could automatically transplant  a feature from one program to another successfully. Another tool, CodeCarbonCopy (CCC), was proposed by Stelios Sidiroglou-Douskos et al.~\cite{Sidiroglou2017} , which automatically transfers code from a donor to a host codebase by utilizing static analysis to identify and eliminate irrelevant functionalities that are not pertinent to the host system.

Inspired by this line of work, we introduce \FOUNDRY (\Cref{sec:approach}), the first software transplantation approach for SPL reengineering. \FOUNDRY is independent of the programming language, and supports SPL's \emph{domain engineering} and \emph{application engineering}~\cite{Clements2001} processes at the code level.
It tackles each SPL reengineering task, easing some and automating others. \FOUNDRY does not eliminate the manual labour of feature identification, but reduces it to the task of annotating the entry points (\ie the interface) of a feature, or its ``organ'' using transplantation nomenclature. \FOUNDRY amortises this manual step across a sequence of transplantations. 
\FOUNDRY  automates feature extraction; to do so, it uses slicing to overapproximate feature dependencies.  It leverages transplantation to automate the variability mechanism and, simultaneously, tackle slice-imprecision.
Key to \FOUNDRY is mapping software transplantation's ``over-organs'', conservative program slices~\cite{Barr2015}, to product line assets, or features. 
Its use of slicing means that \FOUNDRY does not need specially prepared donors; The donor programs can even be unaware that they are participating in an SPL. A product line via ST is composed of multiple over-organs and a ``product base'', a host that contains all features are shared across all products within the product line, so constructing a product entails transplanting a set of organs into a product base.

Because over-organs are conservative, self-contained slices, two organs may share features. For example, in an editor, two different features, like a spell checker or a plugin manager, might share a memory-resident database feature. \FOUNDRY uses clone-aware genetic improvement~\cite{Petke18} to specialise an over-organ to its implantation point and to detect and remove cross organ redundancies (\Cref{sec:approach}).

We realise \FOUNDRY in \prodscalpel, a tool that transplants multiple organs (\ie, a set of interesting features) from donor systems into an emergent product line for codebases written in C.  
%\ls{variability mechanism is used without definition, just below:  it needs to be defined with a citation.} 
\prodscalpel also supports the use of existing variability mechanisms~\cite{Gacek2001} based on \emph{feature toggle}~\cite{Rahman2016} or \emph{preprocessor directives}~\cite{Kastner2008B}. It can surround implanted organs with feature flags, which permit enabling and disabling features, to facilitate its integration into an existing SPL codebase that uses them.

To evaluate \prodscalpel, we conducted two case studies (\Cref{sec:case_studies}) and a controlled experiment (\Cref{sec:experiment}). We first generate products by transplanting features from three real-world systems ---  Kilo\footnote{https://github.com/antirez/kilo}, VI\footnote{http://ex-vi.sourceforge.net/} and CFLOW\footnote{https://www.gnu.org/software/cflow/} --- into two product bases generated from VI and VIM\footnote{https://www.vim.org/}, used as hosts for the target transplantations. Next, we conducted an experiment in which we asked twenty SPL experts to move a feature into a product line. 
We gave them the same inputs as those \prodscalpel requires. In all cases, \prodscalpel outperformed our experiment's participants in the time taken to transplant the feature.
On average, \prodscalpel took 18\% of the time to transplant features from single systems to a product line than the participants who completed the task within the timeout.

Our results (\Cref{sec:result_discussion} and  \Cref{sec:evaluation_results}) show that software transplantation speeds SPL reengineering, by combining features extracted from existing, possibly unrelated, systems.

The main contributions of this paper are: 
\begin{enumerate}
    \item \FOUNDRY, a novel SPL reengineering approach that leverages software transplantation to extract and reuse features from existing codebases to construct a product line, even when those features and their codebases were not built for, or even aware of, the product line.

    \item \FOUNDRY's realisation for C in \prodscalpel, a  tool that transplants multi-file organs and uses clone detection to prevent implanting redundant features;

   \item A rigorous evaluation of \prodscalpel that demonstrates \FOUNDRY's promise. We use \prodscalpel to generate two product lines and two new products, composed of features transplanted from three different real-world codebases. We also show that \prodscalpel migrates features on average 4.8 faster than SPL experts performing the same task.

\end{enumerate}

 All the source code and data needed to reproduce this work are available at the project webpage~\cite{ProjectWebpage}.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec02-motivating_example.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Motivating Example} \label{sec:motivating_example}

 \begin{figure*}[t]
	\centering \includegraphics[width=\textwidth]{images/incremental_ST4.png}
	\caption{Product derivation process using the \FOUNDRY approach. \textit{\prodscalpel transplants three features, in sequence, into the GEdit's product base to derive a new text editor after three iterations of organ transplantation.} }
	\label{fig:incremental_pd}
\end{figure*} 
 %COMMENT: \ls{The case for SPL needs to be elaborated:  we must argue that the various GNOME subprojects share sufficient code to define a product base.  If too many of them are too different, SPL would be inappropriate.}
%\todo{This sentence is problematic:  we shoudl discuss. Done}
%\eb{from Leandro: I've rewritten this sentence by providing details of how GNOME could benefit from using SPLE principles as mass customisation and a common platform.}

The open-source GNOME project\footnote{https://wiki.gnome.org/Projects} encompasses a large portfolio of individual programs evolve as independently as possible from the rest. These programs share features, but because they are separately developed, their constituent features cannot be easily reused across its portfolio to provide mass customization, at least without much manual effort. The combination of mass customisation and a common platform, principles of Software Product Line Engineering (SPLE)~\cite{Pohl2005}, would allows to GNOME team reuses a common base of technology and, at the same time, to bring out products tailored individual customers. Without a common platform and a software development process base on mass customisation, it may be more difficult to the GNOME project provides customized products and effectively manage the commonality and variability of its features.

The GNOME  project is a natural candidate for SPL, but the significant reengineering investment of time and resources have prevented it from adopting SPL. \FOUNDRY is transformative because it can be used to reduce this cost.

We show how the GNOME team could use \prodscalpel to quickly generate a product line. Suppose project collaborators want to build a product line in the domain of text editors. This product line would allow GNOME to produce text editors that augment its current text editor, \emph{GEdit}\footnote{https://wiki.gnome.org/Apps/Gedit}, with additional features. Since they have decided to augment Gedit, GNOME team would select it as the product base, the shared substrate of a product line that, for \FOUNDRY, serves the host for transplanted features.  Assume that the GNOME team targets the following three features (1) $\texttt{side-panel}$, (2) $\texttt{split pane}$, and (3) $\texttt{presentation}$. They then identify two donors from which to transplant these features:  \emph{NEdit}\footnote{https://sourceforge.net/projects/nedit/}, a multi-purpose text editor that is not part of the GNOME portfolio, and \emph{Evince}\footnote{https://wiki.gnome.org/Apps/Evince}, a document viewer for multiple document formats that is part of GNOME, not not an editor.

\prodscalpel iteratively and incrementally re-engineers the codebase for SPL.  In the donor, engineers need to demarcate all feature entry points to transplant; a single annotation is sufficient for \prodscalpel to extract a feature. 
%\todo{remove: To prepare the host, GNOME engineers use \prodscalpel to extract a product base by removing all not desired features. }

To prepare the host, GNOME engineers use \prodscalpel to extract a product base from an existing system by removing all features not shared across all products within the built product line

Then, the engineers must annotate the host to indicate the implantation point for each target feature, or “organ” using transplantation nomenclature. GNOME engineers then run \prodscalpel on these inputs, once per feature, with 
%\ls{This listing is better.  First, the important thing is to show what a user would \emph{actually} type!  Doing so subtly shows that the system is real.  This text is nothing real and, as such, it invalidates the sentence that proceeds it.  Second, why define these switches?  They're self-defining!  What could the switch "donor\_folder" mean other that the path to the donor source code? I am leaving this todo here briefly for you to read it before I revert your change.  I'm trying to avoid churn.  I would like your help with which ones you think are mandatory.  Many should be either defaulted or in a configuration file, not on the command line.  Thinking more about that, maybe all are configuration, not CLI, and we should present a config file here with a minimal CLI invocation.} \eb{I'm with you Earl, all parameters of this command can be considered configuration. I can improve it in the next version of the ProducScalpel or during the journal review process. Do you prefer I remove this command for the moment?}
\begin{quote}
    $\texttt{./prodScalpel \textemdash \textemdash seeds\_file}$: \textit{ The file which contains the seeds for Genetic Programming(GP) algorithm.}

    $\texttt{\textemdash \textemdash donor\_folder}$: \textit{The path to the donor source code.}

    $\texttt{\textemdash \textemdash host\_target}$: \textit{The file in host that contains the insertion point of the transplant.}

    $\texttt{\textemdash \textemdash donor\_target}$: \textit{The file in the donor that contains the core function.}

    $\texttt{\textemdash \textemdash workspace}$: \textit{The path to the workspace of the transplant.}

    $\texttt{\textemdash \textemdash core\_function\_target}$: \textit{The file which contains all feature entry points.}

     $\texttt{\textemdash \textemdash  host\_project}$: \textit{The path to the product base source code.}

   \end{quote} 
%
%\begin{lstlisting}[language=bash]
%prodScalpel --seeds_file <transplantation_directory>/seed-1.in 
%    --compiler_options <transplantation_directory>/CFLAGS
%    --host_target <transplantation_directory>/productBase/insert_point_file.c 
%    --donor_target <donor_directory>/Donor/organ_souce_file.c 
%    --donor_folder <donor_directory>/Donor/
%    --workspace <transplantation_directory>/ 
%    --core_function <core_funtion_name> 
%    --host_project <transplantation_directory>/ProductBase
%\end{lstlisting}
%
%\ls{\lstinline+<root>+ is used without definition.  How can the Donor and Host share the same root?  That undercuts our contribution that the Donor can be unaware of the host!  Some of these switches look like they can be defaults, like seeds and compiler\_options and NOT reported/shown here.}

This command automatically extracts all of the specified feature's source code and its dependencies, or “over-organ” using transplantation nomenclature. More operational parameters are available at the project: webpage~\cite{ProjectWebpage}.

\Cref{fig:incremental_pd} illustrates all transplantation iterations performed to generate a new product.  It shows a new text editor derived from the transplant of features from different donors and using GEdit as a product base. Using feature models~\cite{Kang1990} to represent each donor system and the product base evolution, \prodscalpel first transplants the \emph{side-panel} feature, extracted from GEdit itself. This transplantation demonstrates that \prodscalpel can transplant features into a product base that comes from the same codebase.  Next, \prodscalpel transplants the \emph{split\_pane} feature from NEdit. It shows how \prodscalpel manages to transplant features from distinct codebases, which is not possible without manual effort using the current state-of-art to SPL reengineering. Finally, \prodscalpel transplants the \emph{presentation} feature from GNOME's Evince renderer.

\FOUNDRY facilitates transplanting features from any program into a product line, opening the door to large scale feature reuse. Open-source projects, like GNOME, are an especially promising source of code for \FOUNDRY, so long as the donors and target hosts share compatible licenses.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec03-approach.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Foundry} 
\label{sec:approach}

%\eb{Will resume working from here.}
%\todo{Paulo veja esta frase abaixo. Ao meu ver ela não esta correta.}

Based on software transplantation idea, \FOUNDRY treats \emph{product base} and \emph{over-organs} (representing features) as product line assets. 
A product base is a host that contains all features that will be shared among the products. 
An over-organ, in turn, is a completely functional and reusable portion of code extracted from a donor system that conservatively over-approximates the target organ~\cite{Barr2015}. An over-organ can be specialized to became an organ that preserves the original behavior of the feature in a different host codebase~\cite{Barr2015}.

Conceptually, in \FOUNDRY, while the product base provides commonalities (i.e., common features) to the target product line, the variability (i.e., variant features) are provided by the transplantation process, as illustrated in \Cref{fig:product_for_transplantation}. This idea opens new ways for SPLE area by automated construction of different products by transplanting multiple organs into a product base.

\begin{figure}[t]
	\centering \includegraphics[width=\textwidth]{images/product_for_transplantation7.png}
	\centering 
	\caption{An overview of how new products are derived from a product line based on ST.}
	\label{fig:product_for_transplantation}
\end{figure}

%\todo{to be removed: It is important to note that \FOUNDRY offers two ways of creating products. One can store over-organs into a given platform of core assets (as in traditional SPL, where a selection of features is enabled). Alternatively, we can create a product by directly extracting and transplanting an organ from a donor into a product base, even if the over-organ is not present in our transplantation platform.}

It is important to note that \FOUNDRY offers two ways of creating products by transplanting organs from a pre-established/created \emph{transplantation platform}, a repository of transplantation assets, or by directly extracting and transplanting an organ from a donor into a product base, even if the over-organ is not present in the transplantation platform. The two ways can also be combined to create specialized products.

In the rest of this section, we overview \FOUNDRY's workflow, describing how it applies software transplantation idea to reengineering of product lines from existing systems. \FOUNDRY is independent of the programming language, and supports SPL's \emph{domain engineering} and \emph{application engineering}~\cite{Clements2001} processes at the code level.

\subsection{Domain Engineering}

In SPL's lifecycle the domain engineering corresponds to the process of establishing a reusable platform of core assets~\cite{Clements2001}. The process defines what will be shared among the products derived from it, i.e., commonalities. It also specifies the possible variations expressed as artefacts, that will enable the customization of product line applications, i.e., products.

In \FOUNDRY, domain engineering corresponds to the process of establishing a product line composed of a product base and a set of reusable over-organs extracted, both stored in the transplantation platform. \Cref{fig:foundry_dom} illustrates the stages of the domain engineering process.

As in medicine, \FOUNDRY has a \emph{preoperative} stage where donors and the host are prepared for the transplantation process and a \emph{postoperative} stage where we evaluate if the transplantation was successful (\Cref{sec:application_eng} for more details on the postoperative stage). The preoperative stage defines pre-transplantation tasks, responsible for the variability analysis process, the organ's test suite, donor and host preparation.

\begin{figure}[t]
	\centering  \includegraphics[width=\linewidth]{images/FOUNDRY_DOM6.png}
	\centering \caption{Domain engineering process supported by FOUNDRY. \textit{Four over-organs (A, D, G, L) are extracted from three donor systems and kept in the transplantation platform, with product base consisting of 2 features shared across all products (P and Q).} }
	\label{fig:foundry_dom}
\end{figure}

%\subsection{Preoperative stage}

\textbf{Variability Analysis.} The preoperative stage starts with a variability analysis process to discover features in existing products with the potential to create the target product line. This process aims to create a variability model to express the valid combinations of features among the donor systems from which they will be extracted.

A variability model can be represented by using a feature model~\cite{Kang1990}. We augmented the traditional feature model representation to incorporate software transplantation inputs required to manipulate and maintain the over-organs in the product line. Each over-organ representation in the feature model is annotated with its corresponding entry point in the donor codebase. 
An organ's entry point is a function in the donor system that belongs to the organ, defines an execution environment expected for its initialization, and provides access to the organ's test suite~\cite{Barr2015}. To determine the organ’s entry point, the SPL engineer needs to provide the name of the function implemented in the donor codebase.

\textbf{Donor Preparation.} This task in the preoperative process consists of cleaning up of donor codebases.  
These can contain some code that will never be used in the target products.
For example, codebases written in C, in general, have code fragments guarded by \#ifdef C-preprocessor directives~\cite{Tartler2011})  commonly used to control code extensions related to features. Although useful for the donor program, such code, if transplanted as part of the target organ, will generate dead code ~\cite{Tartler2011} that will never be executed in any transplanted product.

Previous work~\cite{Barr2015} with focus on transplantation of a single feature did not concern with donor clean-up. 
However, even when transplanting a single organ, dead code, if not removed apriori, can lead to unnessasary bloat and lower efficiency of the over-organ adaptation process (\Cref{sec:application_eng}). 
In \FOUNDRY, the donor clean-up can be performed in a manual or automated way. During the process the source code structure of the program needs to be preserved (indentation, spacing, number formats, etc.), to prevent future bugs.

\textbf{Organ Test Suite Preparation.} For product derivation, an SPL engineer must supply test suites, called \emph{ice-box tests}~\cite{Barr2015}. They are used to guide genetic programming in the over-organ adaptation process to create an organ that is fully executable when implanted in the product base (see \Cref{sec:organ_reduction}). 
Ice-box tests can be easily implemented as proposed by~\cite{Barr2015} and integrated into the transplantation platform to be used for new transplantations of the target organ. 
These can be quick to develop, using existing test generation tools, or even adapted from donor’s unit tests, when available.

Even though the preparation process may require manual effort for identifying features of interest,  localising and removing  dead code from the donor and preparing all test suites for the organ, it can be amortised across multiple transplantations and reuse of a single over-organ.

\textbf{Host preparation.} The SPL engineer has to select a product base. It is an existing system which already provides a set of solutions (features) so close to the target products that it can be used as a baseline for the assembly of products. For example, a text editing program could provide a baseline for new programs for text translation,  presentation or rendering, since they could have a considerable number of common features between them.

In case it is necessary, the product base can be reduced to its basic form, keeping only mandatory or features relevant to the target product line. The reduction process consists of removal of all code that implements all features which will not be required to create the target products. In a removal scenario, an additional attention is necessary to find and remove all portions of code that implement each unwanted feature without compromising the product base.

A product base, once reduced, can be used multiple times as a base for new products. Although its preparation may require considerable effort for localising and removing all unnecessary code, it can be compensated with the benefits achieved through using the product base as a baseline to build other products belonging to the same or similar domain.

\textbf{Over-organ Extraction.} Once the donor is prepared, it is possible to start the over-organ extraction process. In this stage,  all source code related to the organ to be transplanted, that implements the target feature, must be identified and extracted.

Conceptually, an over-organ is composed of an organ and its vein, in keeping with the transplantation analogy~\cite{Barr2015}. While the organ implements the functionality we wish to transplant, the vein is a path from the donor entry that builds and initializes an execution environment for the organ~\cite{Barr2015}. Thus, all the code belonging to the organ and its vein must be captured in the extraction process.

In practice, organs can consist of several lines, one or more classes, as long as they fully implement a specific functionality~\cite{Wang2018}. We consider an organ to be a functional implementation of a feature of interest to an emergent product line. 
This makes the code extracting process a challenging task since it also involves identifying all code elements for the organ to be kept functional even out of the donor's environment.
The extraction of a target over-organ can thus involve a considerable amount of code at different levels of granularity, from moving required files and libraries to entire functions and individual statements, both potentially not confined to a single class, file or library~\cite{Wang2018}. For instance, the feature $\texttt{FEAT\_DIFF}$ implemented in VIM has more than 5k LOCs scattered across 33 of its 166 source files. 
Previous work extracted features from a single file~\cite{Barr2015}.

Here we have four challenges that need special attention. 
First challenge concerns the integrity of the extraction when the extracted code appears in several files, issue also identified by Wang et al.~\cite{Wang2018}. 
In this case, the extracted organ also needs to preserve the original multi-file structure. 
Otherwise, it may be even more difficult to keep it functioning outside of  the donor's environment or to propagate eventual changes to the desirded feature (bug fixes, enhancements, etc.).
Second, defects might be introduced due to code redundancy stemming from multiple organ transplantations. 
Multiple organs can use the same code, creating code duplication and possible errors, if not handled correctly.
Third, an organ's vein can contain a large amount of code, especially when the host and the donor have very different structures. 
This requires extensive modification during the adaptation process. 
According to Barr et al.~\cite{Barr2015},  inlined function calls constitute the second largest number of code lines transplanted. 
Fourth challenge is that an extracted over-organ may itself contain multiple smaller features, which its functionality depends on.
For example, a $\texttt{spell\_checker}$ feature might depend on a memory-resident database feature. 
Thus, the extraction process has to implicitly learn feature's dependencies, by including them in its over-organ. 
Such issues, if not solved, make impracticable the generation and maintenance of product lines using the software transplantation approach.

To solve all additional challenges highlighted above, we have evolved the organ extraction process, introduced by Barr et al.~\cite{Barr2015}, to compute slices in multiple files. 
Thus, even if an over-organ is contained in multiple files, \FOUNDRY can obtain a practical over-organ for transplantation, maintaining its original structure.

Given an entry point in the donor provided by the user, \FOUNDRY uses conservative slicing to automatically extract a feature into an ``over-organ'', completely automating the extraction task. Based on the method introduced in \cite{Barr2015}, \FOUNDRY slices \emph{forward} and \emph{backward} from the given organ entry point to identify the organ and one of its veins.

At the end of the extraction process, the  source code of the over-organ is then stored in the transplantation platform, together with other over-organs that compose the product line. 
All over-organs in the platform are available to be reused during the application engineering process.

\subsection{Application Engineering}
\label{sec:application_eng}

In SPL, \emph{application engineering} corresponds to the phase where features are assembled to create a product. This is the phase where variability is realized so that artefacts customization takes place.

In \FOUNDRY, application engineering corresponds to the phase of developing customized products through the organ transplantation process. 
After the execution of multiple iteractions of organs transplantation a new product is derived as new organs are transplanted. Thus, the flexibility required to customize products is provided by extracted over-organs that are combined with a product base.

As illustrated in \Cref{fig:foundry_app}, application engineering process is supported by \FOUNDRY by applying four stages of software transplantation: (i) over-organ selection, (ii) over-organ reduction and adaptation, (iii) organ implantation and (iv) postoperative stage.

\begin{figure}[t]
	\centering  \includegraphics[width=0.9\textwidth]{images/FOUNDRY_APP5.png}
	\centering \caption{Application engineering process supported by FOUNDRY. \textit{A new product is derived after two software transplantation iterations (organs G and L).}}
	\label{fig:foundry_app}
\end{figure}

\textbf{Over-organ Selection.} At the start of the transplantation process, an SPL engineer selects the target features that will be transplanted into the product base to create the target product. 
The choice is guided by the feature model generated during the variability analysis process, so as, to support the SPL engineer to handle eventual relationship and restrictions among transplanted organs. Once the target feature is chosen in the feature model, it is possible to find the corresponding over-organ in the transplantation platform.

\textbf{Over-organ Reduction and Adaptation.} The over-organ reduction and adaptation processes involve specialising the organ to the host environment, i.e., the product base. An SPL engineer must select the target product base with an annotated implantation point where a call to the organ will be grafted to initialize and execute it.

Barr et al.~\cite{Barr2015} automated the over-organ reduction and adaptation process using a GP algorithm.
GP reduces an over-organ and specialises it to the host environment.
It thus creates an organ that preserves the original behaviour of the feature at a given insertion point in the host environment. 
However, Barr et al.'s approach does not support the adaptation of an organ that contains multiple files. 
Moreover, it does not support organ maintenance tasks, but rather provides a one-off transplantation approach.
To overcome these issues \FOUNDRY introduces an organ-host wrapper.
This layer is responsible for providing access to the organ from the target host. 
It is automatically constructed on demand, according to a given implantation point in the product base.

The organ-host wrapper frees developers from the burden of manually writing the code to convert host's data structures into parameters to the organ's entry point whenever a new product is demanded from the product line.

\label{implantation}
\textbf{Organ Implantation.} In this stage of the process, the organ is ready to be implanted into the product base. 
However, to make the application of this technique in generating SPL feasible, we have to consider the transplantation of multiple organs into a single host, a product base, including ones extracted from the same donor. 
As a consequence, the process is no longer concerned with a single but multiple organs and its consequent dependency and interactions.

Feature dependency is a well-known problem in software reuse~\cite{Ribeiro2011}. Dependencies among features are established by means of structural dependencies in the source code shared between elements of different features~\cite{CafeoA2016}. In practice, an organ implementing a feature in a system often shares elements, such as variables and functions with one or more organs that belong to the same donor. For instance, a structure that stores data that are manipulated by more than one function or file; or a function call between the code that belongs to organ A and B. When this happens, the \emph{organ collision} problem occurs since both organ A and B extracted from the same donor must contain the shared function.

\Cref{fig:organs_connection_point} shows a real-world example of two call graphs from the same donor, GEdit text editor, sharing several functions. If we consider them as part of two unrelated organs, all common functions (highlighted by blue boxes) will be duplicated during their corresponding implantation processes. As a consequence of organ collision, transplantation process can insert code that is duplicated from multiple organ transplantations. Such a problem, if it is not managed, will lead to unwanted duplicated code, possibly breaking the postoperative product.

\begin{figure}[t]
	\centering \includegraphics[width=\linewidth]{images/organs_dependency3.png}	\caption{Call graph extracted from GEdit text editor. \textit{An example of connection points among call graphs from organs $\texttt{copy}$ and $\texttt{search\_next}$. Highlighted with blue boxes are functions belonging to both organs.}}
%	\Description{The 1907 Franklin Model D roadster.}
	\label{fig:organs_connection_point}
\end{figure}

In \FOUNDRY, code elements (functions, directives, constants, declarations of several global variables and their definitions) already belonging to the beneficiary or to more than one organ are characterized as \emph{implicit connection points} since they can represent a connection or dependence points among two or more organs.

To correctly work, the implantation process need to identify all duplicated code elements and insert them only once into the product base, avoiding code duplication. However, hosts tend to have large input spaces into which code is inserted. In this way, finding the implicit connection points in the host can be difficult. For instance, functions can have the same namespace but not be identical. Thus, it is necessary to check whether a specific code element is already present in the host, considering not only its namespace but its structure and context at a fine level of granularity to make sure that two portions of code are "clones".

Such particular aspect represents a new challenge in the software transplantation field for SPL, handled by \FOUNDRY. 
We solve this challenge by using code clone detection, to avoid duplicated code insertion.

\textbf{Postoperative Stage.} As in medicine, \FOUNDRY requires checking the side-effects of the transplantation operation. Based on \cite{Barr2015}, \FOUNDRY's postoperative stage introduces three validation steps, as outlined in \Cref{fig:postoperative_tests}. Extending the validation process proposed in~\cite{Harman2013, Barr2015}, we highlight the three test suites that \FOUNDRY uses to evaluate the quality of a transplant: \emph{Regression}, \emph{Regression++}, and/or \emph{Acceptance} tests.
Once selected, organ has been successfully transplanted, and the postoperative product has passed all postoperative validation steps, we incorporate the Regression++ and Acceptance tests into the host's existing regression test suite for use in the next transplantation.

\begin{figure}[t]
\centering \includegraphics[width=0.6\textwidth]{images/postoperative_tests2.png}
\caption{Three validation steps; the dashed boxes are test cases added into the host regression test suite after each transplant iteration.}
\label{fig:postoperative_tests}
\end{figure}

After the postoperative stage, new iterations of organ transplantation can be performed; thus, in a stepwise and incremental way, a new product is derived as organs are transplanted into a product base.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec04-implementation.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Implementation} 
\label{sec:implementation}

We now present \prodscalpel, our realisation of \FOUNDRY for C. \prodscalpel extends $\mu$Scalpel~\cite{Barr2015} to transplant multiple organs at once and implant them into a single host codebase. This feature is essential for supporting \FOUNDRY, since many features, especially those existing, SPL-oblivious codebases, span multiple files.

%\eb{from leandro: I've write these two next sentences describing the figure 7 while it makes a connection between section 3 and 4.}

The overall architecture of \prodscalpel comprises five distinct modules, as depicted in Figure \ref{fig:prodscalpel}. These modules serve as integral components of the \prodscalpel solution, implementing the different stages of \FOUNDRY for the reengineering of systems into SPL. 
The domain engineering process is automated by the \textit{Preprocessing and cleansing} and the \textit{Over-organ extraction} modules. The first one is responsible for cleansing the donor codebases, eliminating any extraneous preprocessor directives that may be present. It also used to host preparing process by removing undesired features delimited by preprocessor directives from the product base. On the other hand, the \textit{Over-organ extraction} module employs the program slicing technique implemented with SGD to extract an over-organ from the donor codebase, thereby isolating the relevant code fragments for further adaptation.

%This step ensures that the host codebase is streamlined for the subsequent transplantation process.

The application engineering process is automated by the \textit{Over-organ reduction and adaptation}, and the \textit{Organ implantation} modules. The first one utilizes genetic programming (GP) to reduce the size and complexity of the selected over-organ obtained from the transplantation platform. This reduction process is carried out while ensuring that the over-organ is effectively adapted to function within the target product base. GP enables the systematic exploration and optimization of the organ's code structure, achieving an organ specialized to a specific product base. The \textit{Organ implantation}, in turn, employs a clone detector to identify code elements duplication and dependencies during the process of implanting the organ into the product base.

More details on how \prodscalpel automates the \FOUNDRY process is provided in the following sections.

\begin{figure*}[t]
	\centering  \includegraphics[width=\textwidth]{images/PRODSCALPEL_ARCHITECTURE.png}
	\centering \caption{Overall architecture of \prodscalpel. \textit{An SDG is a system dependency graph that the genetic programming phase uses to constrain its search space.}
 }
	\label{fig:prodscalpel}
\end{figure*}

\subsection{Automating Feature Removal}

First we need to select and prepare our product base. The selected program might contain some unwanted features for the target product, thus we want to remove them before transplantation takes place. To automate removing unwanted features and dead code from the donor and host systems, \prodscalpel  implements a  \emph{Reconfigurator}. To work, \prodscalpel initially requires engineers to provide as input a textual list of preprocessor directives corresponding to each feature and annotation to be removed. From this, the tool searches for pieces of code that implement features limited by preprocessor directives, removing them from the product base. \prodscalpel then maps and searches selectively for pieces of code that implement features limited by such directives. Then, \prodscalpel removes all features source code from the product base while it keeps the source code structure belonging to the product base unchanged and ready to receive the transplanted organs. \Cref{fig:reconfigurator} gives a example of a portion of code after \prodscalpel cleaned up unused directives.

\begin{figure}[t]
	\centering \includegraphics[width=0.7\textwidth]{images/reconfigurator2.png}
	\centering 
	\caption{The reconfigurator used to automated removal of preprocessor annotation from donor codebase and unwanted features from product base.}
	\label{fig:reconfigurator}
\end{figure}

\subsection{Automating Vein and Over-Organ Extraction}
As optimisations, \prodscalpel extracts the vein from multiple files. \prodscalpel uses backward slicing to identity each function belonging to the vein and saves it in a copy of its souce file in the transplantation platform. Then, it includes all functions calling existing into vein source code in the \emph{array of over-organ statements} used by GP to achieve an organ from the over-organ~\cite{Barr2015}. Thus,  a function/method in the vein shared between two organs is not duplicated when transplanted.

\prodscalpel avoids the problem of vein redundancy when a great part of the vein is common for multiple organs transplanted from the same donors. The vein duplication is discovered by the implantation module during the implantation stage. In the extraction stage, both veins belonging to two organs (A and B) are kept duplicated in the transplantation platform in different directories, since it is important to keep the over-organ functional. However, when the second organ( organ B) is implanted after the transplant of organ A, the code clone detector looks for duplication also in the vein code of organ A. For example, imagine that a vein has a function $\texttt{fx()}$ implemented in file $\texttt{F\.c}$ and belonging to both over-organs A and B. When \prodscalpel tries to transplant organ B after organ A, it checks if the function  $\texttt{fx()}$ already exists in file F.c in the post-operative environment. In this way, if the function  $\texttt{fx()}$ is already in the host post-operative, as part of organ A, \prodscalpel does not insert it into the host again. Instead, the function $\texttt{fx()}$ already transplanted also is used by organ B.

\prodscalpel also identifies occurrences of mutually recursive functions, even an occurrence of an indirect recursion. Technically, \prodscalpel inlines the vein code while puts each function found in a stack of functions. When \prodscalpel finds an occurrence of a recursive function it recovers the beginning of the recursive call and does not inline it. Instead, \prodscalpel extracts the function to a file with the same name of the where the function is implemented. Then, it inserts a calling from the vein to the recursive function. This solution provides a finite interpretation for not inlining mutually recursive functions in the array of over-organ statements. Thus, \prodscalpel can produce a search space for GP with a finite amount of program statements even from recursive functions.

To handle transplantation of organs spread in multiple files, \prodscalpel records the name of the files where the slice is and its location with respect to other slices from the over-organ. 
Then, it records the related statements in an Abstract Syntax Tree (AST), according to their order of appearance in the file. 
This is done to preserve the same structure in the transplanted organ as it appeared in the donor.
Then, \prodscalpel computes the resulting slices in copies of its original files  in the transplantation platform, without breaking the over-organ.

\subsection{Automating Over-organ Reduction and Adaptation}\label{sec:organ_reduction}
\prodscalpel improves the process of over-organ reduction and adaptation by being able to handle over-organs containing multiple files.
It also introduces a layer in the organ that works as an organ-host wrapper.

In practice, \prodscalpel uses GP, as in previous work~\cite{Barr2015}, to prune one or more program elements within the boundaries of the target organ while keeping the organ still functional and passing on the icebox tests. 
In the wrapper, \prodscalpel~abstracts variable names so that GP can select a type-compatible binding. It selects different combinations of all valid statements, variables and function calls mapped from the organ's vein to initialise an execution environment that the organ expects before executing it.

\prodscalpel uses GP to search for matching between variables in the organ and the product base, during the over-organ adaptation process. The matches found are inserted in the organ-host wrapper. 
By a mutation operation, a new version of the organ (i.e., a new individual) is created while \prodscalpel makes several changes in the organ-host wrapper and pruning the over-organ. Each such mutation operation is either an $\texttt{INSERT}$, $\texttt{REPLACE}$ and $\texttt{DELETE}$ of code into the individual and the wrapper at the level of statements.

In the end, \prodscalpel synthesises a call to the extracted organ to execute and test it from the wrapper constructed.

\subsection{Automating Multiple Organ Implantation}
\label{sec:Implantation}

Once the organ is adapted to correctly work on the host it can be automatically implanted into the product base. To correctly insert an organ into the tool must identify and handle potential implicit connection points by avoiding the insertion of code duplication into the product base.

\Cref{fig:code_clone_analysis} illustrates our solution to avoid the organ collision problem. To sum up, the clone detector checks if a specific code element is already present in the beneficiary's environment. To do this, \prodscalpel constructs two lists, one with elements in the organ and other with elements in the host. Then, it checks if there is some element in the target organ which is already presented in a list of code elements transplanted. Once a potential element duplication is identified, \prodscalpel decomposes both organ and host elements in ASTs to detect semantic dependencies between organ implementations previously transplanted, preventing shared elements to be inserted into the host again. Then, the code element is entirely \emph{grafted}, \emph{discarded}, or \emph{merged}. In this last case, \prodscalpel introduces additional line breaks such that potential variances within statements and other structures can be accurately inserted by using sub-abstract tree comparison.

\begin{figure}[t]
	\centering \includegraphics[width=\textwidth]{images/clone_detector.png}
	\caption{Code clone detector. \textit{It performs two comparison steps: text-line and ASTs comparisons to identify code clones and make decision if a code elements must be graft our not into the target product base environment.}}
	\label{fig:code_clone_analysis}
\end{figure}

We augmented \prodscalpel with a \emph{code clone detector}, based on NiCad~\cite{Roy2009}. This clone detector finds exact clones over arbitrary program fragments in the organ and host source code by using ASTs. Thus, we exploit the benefits of \emph{Program differencing}~\cite{Kernighan1983} technique and TXL~\cite{Cordy2006} to identify and compare potential syntactic code duplication using text-line and ASTs comparison~\cite{Roy2009}.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec05-automated_spl_reengineering.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Automated support for product line re-engeneering} \label{sec:automated_spl_reengineering}

Automated software transplantation, as presented in \FOUNDRY, can provide a set of opportunities to improve and automate the product line reengineering process. This alone might justify the effort to explore research opportunities in software transplantation for SPLE. Here we elaborate on and discuss such opportunities.

\textbf{ Automating reengineering of existing systems into SPL.} Principally \FOUNDRY is proposed as a new way of reengineering existing systems into SPL. 
Companies can use \FOUNDRY just for the initial conversion of an existing codebase into SPL. In this mode, \FOUNDRY first extracts over-organs from the donor systems, then, produces a product base, created from an existing system by removing all unwanted features to the product line. Then, adapts and implants the organs into this product base, identifying and removing cross-over-organ redundancies. The resulting organs become the SPL's shared set of features. \FOUNDRY can itself be used as a variability mechanism to construct new products by implanting features into a product base. Even with this one-off application of \FOUNDRY, a company can also use \FOUNDRY to surround the transplanted feature with  a conventional variability mechanism. For existing SPL codebase, such as the one it created. \FOUNDRY's variability mechanism insertion also supports surrounding implanted organs with feature flags or preprocessor directives, which permit enabling and disabling of features, to facilitate its integration into an existing SPL codebase that uses them.

\textbf{Automating clone-and-own technique.} \FOUNDRY can automate \emph{clone-and-own}~\cite{Dubinsky2013, Fischer2015}, especially the task of synchronising changes to a feature shared across two products created by clone-and-own. For example, consider the case where the copy of a shared feature in one of the two products is patched to fix a bug. \FOUNDRY permits transplantation of the fixed version over the top of the unpatched copy of the feature.

\textbf{Automating Reactive Product Line Adoption Process.} 
Certain companies already have product lines, but adapt a reactive process, where a new feature is added only when a need for such a feature arises. 
\FOUNDRY can be used initially to generate product variants, and then, as the demand for specific products increases, it could be used to generate product lines.

\textbf{Automating a symbiotic SPL.} \FOUNDRY permits a wholly new form of SPL, \emph{symbiotic SPL}:  in this mode, the donor can be oblivious to a parallel, ongoing SPL reorganisation of its codebase. To capture improvements in the donor, \prodscalpel periodically refreshes its set of features by re-transplanting them into its donor product base and hence into host products. For instance, one could use \prodscalpel to produce lightweight, specialised text editors from the \emph{Vim} project.

\textbf{Supporting Controlled Maintenance and evolution of SPL.}
The maintenance of assets and products is a challenging task and an inherent characteristic of a product line. When we include organs as assets in a product line the process becomes even more challenging. In fact, by trying to maintain an over-organ individually in the platform, a developer might introduce errors into the product line or in products derived from it, since their organs eventually share elements —such as variables and functions—with the maintained organ. A solution would be to re-implant the changed organ. Nevertheless, a crucial problem here is how to re-implant the organ changed to match a different version of it already transplanted in the target product.

We envision two ways to avoid this problem and \emph{maintain} a created product line. Firstly, one can re-transplant the features if the original source codebase changes. Secondly, one can maintain the extracted over-organs, and re-run the adaptation and implantation stages, as need be.

In both scenarios of product line maintenance, \FOUNDRY can be set up to evolve a changed organ with feature flags. Surrounding an organ with feature toggles just before implantation is an interesting alternative for continuous deployment of organs. It thus separates inserted code and makes it easier to maintain.
To support continuous deployment, the implantation process in \FOUNDRY can be configured to involve each organ with feature flags to connect new, unreleased code to production. Once a transplanted organ is ready for production, developers can turn off the flag and reveal the new organ or its changes to users.

\textbf{Providing a Variability Mechanism through Organ Transplantation.} 
\FOUNDRY provides a variability mechanism itself. It allows developers to include or not a given feature, to instantiate different products at any moment only by transplanting them into the target product. It avoids the problem of maintaining and evolving a product line that consists of a massive number of products. Additionally,  it avoids feature flag removal and technical debt around unremoved feature flags.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec06-case_studies.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Case Studies} \label{sec:case_studies}
We evaluate \prodscalpel~on two case studies for generating new product variants from real-world systems. This section explains the objectives, research questions, and subject systems. Additionally, it presents the results of our study and discusses the implications. Our corpus and data collected are available at the project webpage~\cite{ProjectWebpage}.

\subsection{Objective and Research Questions}
The objective of this study is to evaluate the proposed approach and tool, thereby demonstrating the potential of automated software transplantation for product line generation. % and 
We aim to answer the following research questions:

\begin{quote}
\textbf{RQ1.} \emph{How much effort is required to generate products from a product line created using \prodscalpel?}
\end{quote}
Given the complexity of the task at hand, it seems unreasonable to expect the product derivation process to be instantaneous. Still, it will need to be fast enough to be incorporated into a development cycle and demonstrate its advantage over manual effort. To answer this question, we computed the time required, and the number LOCs transferred to productize software.

\begin{quote}
\textbf{RQ2.} \emph{Which features can \prodscalpel~productize?}
\end{quote}
We discuss the characteristics of features that \prodscalpel~can and those it cannot yet handle. Given the inherent complexity of the process, it is important to clearly define its current limitations.

\subsection{Methodology}

\textbf{Subject systems}. We chose two text editors, \emph{VI} and \emph{VIM}, as product bases.
For donors we chose subjects from two different domains: code analysis software and text editors.
We chose \emph{GNU Cflow}, a call graph extractor from C source code, as a donor.
This is to show that donors can also come from different application domains.
We reuse \emph{VI} and \emph{VIM} as donors, as well as another text editor, \emph{kilo}.
This is to show that donors can come from the same system as the product base, but also different systems within the same application domain.

Our donors are presented in Table~\ref{tab:donors_list}.
We identified the following features as possible desired features in a new editor: $\texttt{output}$ from CFLOW, $\texttt{enableRawMode}$ from kilo, $\texttt{vclear}$ from VI, and $\texttt{spell\_check}$ and $\texttt{search}$ from VIM. Table~\ref{tab:donors_list} presents more details on the subjects.

\textbf{Procedure and execution}. 
Initially, we automatically remove dead-code from both donors and host codebases by using \prodscalpel. 
We also reduced each host to its basic form removing all optional features. Thus, both donors and host are prepared for the transplantation process. 
Given an organs' entry point for each organ in the donor systems provided by us, their target implantation points in the product base, and a set of test suites for each organ, \prodscalpel was used to localise and extract a set of organs from the donor, transform each organ to be compatible with the context of their target sites in the product base and implant the organs in the beneficiary's environment. Each automated organ transplantation process was repeated 20 times, due to the heuristic nature of the over-organ adaptation process. We computed the average number of lines of code transplanted and the average runtimes for the transplantation process.

\textbf{Case study environment}. The runtimes for each transplantation were measured on an Intel Core 3.1 GHz Dual-Core Intel Core i5, with 16 GB memory running MacOS 10.15.4.

\begin{table}[t]
\centering 
	\caption{Donors and hosts corpus for the evaluation: column Features shows the number of features identified. 
	}
	\label{tab:donors_list}
	\begin{tabular}{llrr} \hline
        Subjects &Type  & Size (LOC) & \#Features   \\\hline
	    Kilo     &Donor & 804        & 17      \\
		CFLOW    &Donor & 4,274      & 54      \\
		VI       &Donor & 20,292     & 36      \\
		VIM      &Donor & 839,438    & 176      \\\hline
		VI       &Product Base & 20,223 & 36      \\
		VIM      &Product Base & 737,466 & 117     \\\hline
	\end{tabular}
\end{table}

\subsection{Results and Discussion} \label{sec:result_discussion}

Table~\ref{tab:transplantation_time} shows average runtimes of transplanting each organ into the product base as well as the total number of code lines transplanted to derive each product. At the end of the transplantation process, the postoperative product A has a total of 28k LoC and 40 features while product B has a total of 745k LOC and 121 features. Together, donors provided three feature variants to the product line, approximately 7.8k LOC to product A and 8.1k to product B, including a feature removed and re-transplanted in \emph{VI}.

\begin{table}[h]\centering

%	\resizebox{12cm}{!}{%
    \caption{Case studies results} 
    \Description{Multi-organ transplantation results to generate products A and B. \emph{Columns Trans. Time shows the time (Sys+User) spent on the organ transplant process in which column Sys. correspond to the \prodscalpel's execution time; column User correspond to the user time spent in preoperative stage and  augmenting the host’s regression test suite (regression++). Product A uses a reduced VI editor as a product base, while Product B was derived by transplanting features from the donors into the reduced VIM editor.}}
    \label{tab:transplantation_time}
	\begin{center}
	\begin{tabular}{lrrrrrr} \hline
		\multicolumn{1}{c}{Donors} & \multicolumn{3}{c}{Number of}   & &\multicolumn{2}{c}{Trans.time(min)} \\
		\cline{2-4} \cline{6-7}
		 & \multicolumn{1}{c}{LoC}  & \multicolumn{1}{c}{Functions} & \multicolumn{1}{c}{Files} & & \multicolumn{1}{c}{Sys.} & \multicolumn{1}{c}{User}\\\hline
	%	MYTAR       &4 & \sim197 & 8 & 8 &  & 74 & 82\\
		Kilo        & $\sim$963 & 35 & 4 & & $\sim$86 & 32\\
		CFLOW       & $\sim$4,822 & 37 & 8 &  & $\sim$344 &123 \\
		VI          & $\sim$1,983 & 5 & 15 &  & $\sim$1,234 &184\\\hline
		\rowcolor[gray]{.9} Product A  &$\sim$7,768 & 77 & 27 & & $\sim$1,664 & 339 \\\hline
	%	MYTAR       &4 & \sim215 & 8 & 8 & & 74 & 82\\
		Kilo        & $\sim$981 & 35 & 4 & & $\sim$94 & 32\\
		CFLOW       & $\sim$4,898 & 37 & 8 &  & $\sim$428 & 123\\
		VI          & $\sim$2,234 & 5 & 15 &  & $\sim$1,294 & 184\\\hline
		\rowcolor[gray]{.9} Product B  & $\sim$8,113 & 77 & 27 & & $\sim$1,890 &532\\\hline
	\end{tabular}
	\end{center}

    %}	
\end{table}

\begin{framed}
\noindent To answer RQ1, on average, \prodscalpel spent 4h31min/1KLoC for transplanting three features into VI, and 4h40min/1KLOC for transplanting the same three features into VIM.
\end{framed}

To answer question \textbf{RQ2}, we also tried to transplant two more features:  $\texttt{spell\_check}$ and $\texttt{search}$,  both containing larger amounts of code, about 104 KLOCs and 153 KLOCs, respectively.

\prodscalpel uses Doxygen~\cite{Doxygen2018}, a source code documentation generator, to generate the call and caller graphs. 
It thus inherits the limitation of generating an imprecise call graph when dealing with function pointers. 
This leads to unneccessary instructions to be copied, while others missing.
Although possible with an additional manual effort, such limitations made our tool unable to automatically extract the larger organs from VIM. 
We can, in the future, turn to \emph{Dynamic analysis}~\cite{Cornelissen2009} and other code manipulating tools to optimise the efficiency of the slicing process, to efficiently identify those statements in the program slice which have influence on the target organ. Investigating more precise techniques is future work.

\begin{framed}
\noindent To answer RQ2, \prodscalpel was able to productise three features from three different systems. However, transplantation of organs from a larger donor exposed some limitations of the technologies used by our tool.
\end{framed}

In summary, these case studies show initial evidence that software transplantation can be successfully used for building product variants automatically by combining features transplanted from real-world systems.
\prodscalpel inherits limitations of the underlying slicing tools, thus its generalisability can be further improved with progress in that domain.
New studies may give more evidence that our approach is also extensible and flexible in other domains, opening perspectives for future work.\looseness=-1

\subsection{Threats to Validity} \label{ch-6:threats}

The relatively small number and diversity of systems used for transplantation pose an external threat to validity. However, we tried to mitigate it by constructing a possible real-world scenario, i.e., transplantation of features that would be useful in production of new text editors.

In terms of internal threats, we are limited by abilities of the tools we use, in particular, Doxygen, for call graph construction, and TXL for program transformation. We also use testing as a means of validating our approach, which cannot provide a formal proof of its correctness. However, testing is a standard approach in evaluating code in real-world scenarios due to its high scalability.

Additionally, our case studies are limited to a specific implementation technique and specific code-transplantation scenarios; generalization to object-oriented languages and others requires further investigation.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec07-experiment.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Experimental Evaluation} \label{sec:experiment}

%\ls{It's unclear whether participants are creating products or product lines.}

In the previous case studies, we showed the viability of our software transplantation approach, implemented in \prodscalpel, to generate customized products from the existing codebase. Although the validation of the approach is based on empirical evidence, it  is  still important to test its efficiency by comparing it with other tools used to product line migration. Unfortunately, tool support for the reengineering process is limited, in general, they give support for specific activities, such as feature location, refactoring or quality assurance~\cite{Assuncao2017,Kruger2020}. To the best of our knowledge, there is currently no comparable tool that manages to automatically transplant features from distinct donor systems to generate a product line.
Therefore, we compare our approach with current state-of-the-art, namely human effort.
We conducted an experiment that reflects a real-world process of product line migration from existing codebases~\cite{Krueger2002}. In this section, we state our research objectives and describe in detail the experimental setting.\looseness=-1
 Our corpus and data collected are available at  the project webpage~\cite{ProjectWebpage}.

\subsection{Goal}
The goal of this experiment is to analyse the effectiveness and efficiency of our approach compared with the manual process of generating a product line from existing systems, performed by SPL  experts. In accordance with the guidelines for reporting software engineering experiments presented in ~\cite{Wohlin2012b}, we have framed our research objectives using the Goal Question Metric (GQM) method suggested by Basili~\cite{Basili1994}. Our goal is to:

\textbf{Analyse} a software transplantation approach to derive product variants 
\textbf{for the purpose of} comparison 
\textbf{with respect to} effectiveness and efficiency 
\textbf{from the point of view of} the researcher
\textbf{in the context of} an SPL project of product line migration from real-world systems.

\subsection{Research Questions}

In order to achieve the stated goal, we defined two quantitative research questions. These are related to the data collected during the period that the experiment was conducted. The questions are described as follows:

\begin{quote}
\textbf{RQ3.} \emph{How successful is \prodscalpel at feature migration when compared with human effort?}
\end{quote}
We would like to understand how well our approach automatically transfers all required code so that the target feature can run in an emergent product line and compare it with the manual process.

\begin{quote}
\textbf{RQ4.} \emph{How much feature migration time can be gained using \prodscalpel compared to the manual process?} 
\end{quote}
With this question, we evaluate the time spent by SPL experts to \emph{extract}, \emph{adapt} and \emph{merge} features to derive new product variants and compare these with the same processes when using \prodscalpel.

\subsection{Measures}
With the objective to answer our research questions, we defined the measures that must be computed. For each question, one measure was defined. These are described as follows:

\textbf{M1.} For RQ3, the accuracy of our approach is computed by verifying if \prodscalpel successfully migrated new functionalities to a product line by passing all the regression, augmented regression and acceptance test suites. This will check whether or not the output of the transplanted feature is correct with respect to the given test suites.

\textbf{M2.} For RQ4, we simply report the time that is spent on each activity to transfer the target features. These activities are: \emph{code extraction}, \emph{adaptation}, and \emph{merging}. The time for each of these activities was collected individually.

\subsection{Methodology}

We use two donor systems in our experiment:
\emph{NEATVI}\footnote{https://github.com/aligrudi/neatvi}, text editor extended from VI for editing bidirectional UTF-8 text and \emph{Mytar}\footnote{https://github.com/spektom/mytar}, an archive manager.  
Table~\ref{tab:instrumentation} gives more details about the systems used in this experiment. Having in mind manually inspecting a codebase to transfer a feature to a product line is hard, slow, and tedious~\cite{Mahmood2021}, we chose to select small systems to avoid participants getting tired. These codebases were available to download together with a script to automate setup of the environment.

\begin{table}[ht]
	\caption{Details of donors and product base systems used in our studdy.}
	\label{tab:instrumentation}
    \begin{center}
    	\begin{tabular}{l|lr|lr|lr} \hline
    		\multicolumn{1}{c}{Scenario} & \multicolumn{1}{|c}{Donors} & \multicolumn{1}{c|}{LoC}   & \multicolumn{1}{c}{Target features}       & \multicolumn{1}{c|}{LoC}  & \multicolumn{1}{c}{Host}& \multicolumn{1}{c}{LoC} \\\hline
    		I       & NEATVI & 5,276 & $\texttt{DIR\_INIT}$  & 239&\multirow{2}{*}{Product base}&\multirow{2}{*}{5,285} \\
    		II      & Mytar  & 1,046 & $\texttt{WRITE\_ARCHIVE}$ & 170&& \\\hline
    	\end{tabular}
    \end{center}
\end{table}

We recruited 20 SPL experts for the experiment that were divided into two different groups.  We chose to allow participants to use their own work environment by avoiding adaptation bias to a strange environment with the use of unknown tools.
Guidelines provided to the participants of \emph{Group A}\footnote{The guideline for perform the transplantation in  scenario I is available at \url{https://rb.gy/vydhbb}} and \emph{Group B}\footnote{The guideline for perform the transplantation in scenario II is available at \url{https://rb.gy/covz62}} consisted of a process description and systems documentation. Additionally, we provided training on reengineering software systems to SPL to ensure that all participants understood the experiment's objectives.

We used timesheets to record the effort spent on the necessary activities required to transfer features from an existing codebase to a product base, as previously mentioned. In addition to the timesheets, we used two forms to collect information about participants' experience. This data is shown in Table~\ref{tab:participants}. 
We also conducted a post-survey to better understand participants' problems encountered.

\normalsize\subsection{Experimental design}

We answer our research questions by simulating a real reengineering process where two features must be transferred to a product line built over a product base. The experimental design was inspired by documented real product-line migration scenarios~\cite{Laguna2013, Assuncao2017}.

In scenario I, we gathered a group of 10 SPL experts (called Group A) where each one of them had to manually re-transplant all portions of code that implement the feature $\texttt{dir\_init}$ to the product base. We removed this feature from the original version of \emph{NEATVI} to generate the product base used in this scenario. In scenario II, another group of 10 SPL experts (called Group B) tried to insert the feature $\texttt{write\_archive}$ from \emph{MYTAR} into the original version of NEATVI used as the product base. Here, we chose not to provide the post-operative product base used in the scenario I. Instead, we provide the original version of NEATVI, already with the feature $\texttt{dir\_init}$, to prevent possible code errors introduced in scenario I. As we analyse both scenarios separately, we believe that this strategy has minimized possible human bias.

The idea is to use each scenario to represent real scenarios of reengineering to SPL where system variants came from both similar (scenario I) and distinct codebases (an archive manager providing features to a text editor's product line - scenario II) as Figure~\ref{fig:experiment_design} shows.

\begin{figure}[t]
	\centering \includegraphics[width=0.8\textwidth]{images/experiment_design8.png}
	\centering 
	\caption{Experimental design: one factor with two treatments applied in two reengineering scenarios.}
	\label{fig:experiment_design}
\end{figure}

In both scenarios participants are given the same inputs as \prodscalpel.
Thus the independent variable in our experiment is whether the feature migration process is automated or not.

In this experiment, the dependent variables are: success of the product line generation process and the payoff for using our approach. That is, to analyze the success of our approach (RQ3), we evaluate how often the transplanted features pass all the provided test cases.
To analyze the performance of the approach (RQ4), we measured the time spent by participants to extract, adapt and merge one feature into a product base in comparison with \prodscalpel's time to complete the same tasks.

\subsection{Pilot Study}

First, we conducted two pilot studies with 6 graduate students. 
We used the pilot study results to determine the amount of time needed to execute our tasks and the suitable size of features. This allowed us to estimate and plan the number of participants we needed for the main study. 
The pilot study also allowed us to assess whether the participants could properly understand the subject systems and the tasks they should perform.

\subsection{Participants}

After the pilot study, we recruited 20 new participants: 2 undergrads (Un), 9 masters (M), 7 PhD. (PhD), and 2 Post-Ph.D. (Post-PhD). Most of them have more than 5 years of SPL experience and 10 years of software development. The participants are from ten different universities (from U1to U10) and the analysts/developers work in four different companies (C1, C2, C3, C4 and C5). To recruit them, we sent emails to professors from two universities, from different software reuse research groups, to suggest current and ex-members.

Before the experiment, we asked them to answer an online survey, which we used to collect background data about their experience, mainly in software development and SPL\footnote{\url{https://rb.gy/ant4g8}}. 
We created balanced groups (A and B) of participants for each product line generation scenario, based on their experience. 
Table~\ref{tab:participants} shows the details of the participants involved in the experiment.

\begin{table}[t]
\centering 
	\caption{\textit{Details of participants' expertise (in years) and division into groups. Group A worked on scenario I, transplanting a feature from different versions of the same donor system as the host. Group B worked on scenario II,transplanting a feature from a donor system different to the host one}}
	\label{tab:participants}
	\resizebox{7cm}{!}{%
	\begin{tabular}{clllll} \hline
%	\toprule
		Group &Part. &Degree & Inst. & \multicolumn{2}{c}{Exp. (years)}\\ \cline{5-6}
		              &         &  &   & Dev.      &  SPL     \\\hline
		A       &P1  & MSc     & U7   & [1,5)     &  [5,10)  \\%magno
    	        &P2  & Un      & C5   & [10)      &  [1,5)   \\%mateus
		        &P3  & PhD     & U4   & [10)      &  [10)    \\%jonatas
		        &P4  & MSc     & U4   & [10)      &  [1,5)   \\%marco
		        &P5  & MSc     & U4   & [10)      &  [5,10)  \\%Djan
		        &P6  & MSc     & U4   & [10)      &  [5,10)  \\%Renata
		        &P7  & PhD     & U8   & [10)      &  [10)    \\%Alcemir
		        &P8  & PhD     & U9   & [10)      &  [1,5)   \\%Tiago
	            &P9  & PhD & U10  & [10)      &  [10)    \\%Larissa
	            &P10  & PhD     & U4   & [1,5)     &  [1,5)   \\\hline%Stefani

		B       &P11   & MSc     & C1   & [10)      &  [1,5)   \\%Daniel
	            &P12   & MSc     & C2   & [1,5)     &  [1,5)   \\%Rose
		        &P13   & Un      & C3   & [10)      &  [1,5)   \\%Taijara
		        &P14   & PhD     & U1   & [10)      &  [10)    \\%Michele
		        &P15   & PhD     & U2   & [10)      &  [10)    \\%Paulo
		        &P16   & PhD     & U3   & [10)      &  [5,10)  \\%Tassio 
		        &P17   & MSc     & U4   & [5,10)    &  [5,10)  \\%Rafael X
		        &P18   & MSc     & C4   & [5,10)    &  [5,10)  \\%Anna
		        &P19   & PhD     & U5   & [10)      &  [10)    \\%Iuri
		        &P20  & MSc     & U6   & [5,10)    &  [1,5)   \\\hline%Loreno

	\end{tabular}
	}
\end{table}

\subsection{Operation}

Before the participants receive their tasks, we introduced the experiment with a \emph{tutorial} on reengineering of existing systems into SPL. The tutorial took 30 minutes on average.

We provided the participants with the same input as the one required for \FOUNDRY, namely: feature entry points in the donor, the donor’s source code, and a prepared product base with the target insertion point.
We also supplied the participants with ice-box tests that could be used to guide the search for organ code modifications required to check if it continues to be executable when deployed in the host. Additionally, they received a few-sentence description of each feature in the target system and the system’s documentation with donor and host feature models.

The direct costs of this experiment are related solely to the time spent by the researcher with the setup of the experiment itself. 
This involved: specifying the respective annotations for the entry point and insertion points of the features, which took approximately 13 minutes of work for scenario I and 17 minutes for scenario II; creating the test cases, necessary to validate the target features, taking approximately 16 minutes of programming activities; preparing the product base, which took approximately 14 minutes; then, approximately 34 minutes were spent creating all documentation of donor systems including the product base feature model.

To extend the emergent product line with a new feature transferred from the correspondent donor system, all participants were required to perform three activities based on the migration of system variants to the target product line~\cite{Krueger2001,Mahmood2021}: feature \emph{extraction}, \emph{adaptation} and \emph{merging}, with descriptions and instructions provided for each task.

Initially, the participants had to identify and extract all code associated with the feature of interest to a temporary directory. 
Then, each participant had to adapt the extracted feature so it executes correctly in the product base environment, passing all unit tests. In practice, each participant had to change the feature source code to be compatible with the name space and context of its target insertion point in the product base. Finally, it had to insert the feature's code into the product base, and validate its correctness via regression testing.

\subsection{Data collection}

We have provided a task and time registration worksheet. While participants were conducting the assigned tasks, we asked them to take notes of which strategies were being used for each stage of the feature transfer process and why they are performing each specific task. It allowed us to capture strategies and performance data simultaneously.

We have complemented the above setup with a post-survey. This way we can better understand participants' problems and differences between the manual and automated process in both scenarios. We have triangulated the data generated from the experiment with the responses we obtain from the pre and post-surveys.

To establish the time for feature transplantation using our automated approach, we ran \prodscalpel 20 times, and measure the average time spent on feature migration in each scenario. This average time was compared with the time spent by our participants on the manual reengineering process.

Based on our pilot study, we set a time limit of 4 hours for each manual and automated process. 
This is to allow for enough time to transfer required amount of LoC while avoiding participants getting bored or tired.

\subsection{Data Analysis}
\label{sec:experiment_data_analysis}
We used 22 pre-existing regression test suites designed by the \emph{NEATVI} developers to assess the success of \prodscalpel and manual transplantations and answer our \textbf{RQ3}. 
However, these were not designed to test \emph{NEATVI} as a product base with new variants and may not be sufficiently rigorous to find regression faults introduced by the reengineering process. 
To achieve better product line coverage, we manually augmented the host’s regression test suites with additional tests, our augmented regression suites.
Furthermore, we implemented an acceptance test suite for evaluating the transferred feature in both scenario I and II.
We have a total of 30 such tests in scenario I and 33 in scenario II. 
Our test suites provided statement coverage of 72.5\% to the post-operative product line in scenario I and 73.3\% to the post-operative product line in scenario II.

\subsection{Results and Discussion}
\label{sec:evaluation_results}

We summarise our results in Table~\ref{tab:transplantation_results}.  
We report the status of the product base and feature inserted by the participants, the time spent and the number of passing tests for the regression augmented regression and acceptance test suites \footnote{The time measured for the participants is available at \url{rb.gy/ant4g8}}. 
In the first scenario, only one of the participants was not able to finish the process before the timeout. On the other hand, half of the participants were able to finish the process before achieving the timeout in the second scenario and only three of them have been able to insert the target feature without breaking the product base.

\begin{table}[t]
\centering 
    \caption{\textbf{Scenario I:} Donor: NEATVI - Product Base: NEATVI without the desired feature. \textit{Experiment results comparing the time of tool over 20 repetitions with the participants:column product line status shows the generated product line status by participants and tool; column \emph{Execution Time} shows the time spent on the feature transplant by the participants and the average time of 20 runs of \prodscalpel. We highlight the execution time of the participant that moqt quickly completed the task. Columns in \emph{Test Suites} show results for each test suite and report statement coverage (\%) for the postoperative host and for the organ. Columns marked with \emph{PASSED} report the number of passing tests.}}
	\label{tab:transplantation_results}
	\begin{tabular}{lcrrrr}\\\hline
		\multicolumn{1}{c}{}       & & & \multicolumn{3}{c}{Test Suites} \\            
		\cline{4-6}  
		 \multicolumn{1}{c}{Participants} & \multicolumn{1}{c}{ Product Line} & \multicolumn{1}{c}{Execution}& \multicolumn{1}{c}{Regre. (59\%)} & \multicolumn{1}{c}{Regre.++ (70.1\%)} & \multicolumn{1}{c}{Accept. (72.5\%)} \\
		\multicolumn{1}{c}{} & \multicolumn{1}{c}{Status} & \multicolumn{1}{c}{Time (minutes)} & \multicolumn{1}{c}{PASSED}  & \multicolumn{1}{c}{PASSED}  & \multicolumn{1}{c}{PASSED} \\\hline

		 P1 &\multicolumn{1}{c}{OK}  &\textbf{82}  & 22/22 &30/30 &3/3 \\
		     P2 &OK      &\textbf{88}  & 22/22  &30/30  &3/3\\
		    P3 &OK     &\textbf{77}  & 22/22 &30/30   &3/3\\
		     P4 &OK      &\cellcolor[gray]{.9}\textbf{68}  & 22/22  &30/30 &3/3\\
		     P5 &OK      &\textbf{81}  & 22/22  &30/30  &3/3\\
		     P6 &Broken & \multicolumn{1}{c}{\textbf{Timeout}} & 0/22     &0/30     &0/3 \\
		     P7 &OK      &\textbf{87}  & 22/22 &30/30   &3/3  \\
		     P8 &OK     &\textbf{83}  & 22/22 &30/30   &3/3 \\
		     P9 &OK      &\textbf{73}  & 22/22 &30/30  &3/3\\
		     P10 &OK     &\textbf{113} & 22/22  &30/30  &3/3\\
		 \hline 
		 \rowcolor[gray]{.9} \textbf{\prodscalpel} &\textbf{OK in 20/20 runs} &\textbf{20} &\textbf{22/22}   &\textbf{30/30}   &\textbf{3/3} \\
		 \hline
	\end{tabular}
    \bigskip
    \bigskip
    \bigskip
    \caption{\textbf{Scenario II:} Donor: Mytar - Product Base: NEATVI. All other columns are the same as in the previous table.}\label{tab:B}
	\begin{tabular}{lcrrrr}\\\hline
	\multicolumn{1}{c}{}       & & & \multicolumn{3}{c}{Test Suites} \\                       
		\cline{4-6}  
		\multicolumn{1}{c}{Participants} & \multicolumn{1}{c}{ Product Line} & \multicolumn{1}{c}{Execution}& \multicolumn{1}{c}{Regre. (70.1\%)} & \multicolumn{1}{c}{Regre.++ (71.9\%)} & \multicolumn{1}{c}{Accept. (73.3\%)} \\
		\multicolumn{1}{c}{} & \multicolumn{1}{c}{Status} & \multicolumn{1}{c}{Time (minutes)} & \multicolumn{1}{c}{PASSED}  & \multicolumn{1}{c}{PASSED}  & \multicolumn{1}{c}{PASSED}  \\\hline
		 P11 &Broken & \multicolumn{1}{c}{\textbf{Timeout}} & 0/22 & 0/33 & 0/2\\
		 P12 &Broken  & \multicolumn{1}{c}{\textbf{Timeout}} & 0/22 & 0/33 & 0/2\\
		 P13 &Error    &\textbf{181} & 0/22 & 0/33 & 0/2\\
		 P14  &Broken &\multicolumn{1}{c}{\textbf{Timeout}} & 0/22 & 0/33 & 0/2\\
		 P15 &Broken  &\multicolumn{1}{c}{\textbf{Timeout}}  & 0/22 & 0/33 & 0/2\\
		 P16 &Error    &\textbf{114} & 0/52 & 0/33 & 0/2\\
		 P17 &OK       &\cellcolor[gray]{.9}\textbf{104} & 22/22 & 33/33 & 2/2\\
		 P18 &OK       &\textbf{194} & 22/22 & 33/33 & 2/2\\
		 P19 &OK       &\textbf{131} & 22/22 & 33/33 & 2/2\\
		 P20 &Broken & \multicolumn{1}{c}{\textbf{Timeout}} & 0/22 & 0/33 & 0/2\\
		 \hline 
		 \rowcolor[gray]{.9} \textbf{\prodscalpel} &\textbf{OK in 19/20 runs} &\textbf{27} & \textbf{22/22} &\textbf{33/33}  &\textbf{2/2} \\\hline
	\end{tabular}
%}
\end{table}

For each scenario, we also report the number of \prodscalpel runs in which the product derived passed all test cases. For each scenario, we repeat each run 20 times. The success rate was retained for both scenarios I and II, where only one run timed out and the product line generated passed all tests from all test suites.

The results show success rate was retained in both scenario I and II, where we lost only one successful run in the timeout and all products derived passed all tests from all test suites.  
With regards to the manual process, nine participants have successfully transplanted the target feature in scenario I. 
In scenario II, seven of ten participants broke the product base when trying to transplant the target feature, while half of the participants were not able to transplant the feature within the timeout.

\begin{framed}
\noindent \textbf{RQ3:} Our results show that \prodscalpel outperformed participants in both scenarios with only one run timing out.  
Eight of twenty participants (considering both scenarios) were not able to transplant the target features without breaking tests.
\end{framed}

As stated in the definition of measure \textbf{M2} and to answer \textbf{RQ4}, we evaluate the payoff of \prodscalpel. Figure~\ref{fig:experiment_result_time_II} graphically shows the time spent on each activity performed in the reengineering to SPL process. In summary, Group A transferred the target feature from \emph{NETVI} to the product base in 1h24 minutes on average. \prodscalpel turned out to be quicker, successfully transplanting this feature in all 20 trials, taking an average of 20 minutes.

\begin{figure}[t]
	\centering \includegraphics[width=0.9\textwidth]{images/experiment_result-SC1-4.png}
	\label{fig:experiment_result_time_I}
\end{figure} 
\begin{figure}[t]
	\centering \includegraphics[width=0.9\textwidth]{images/experiment_result-SC2-4.png}
	\centering \caption{Time (in minutes) spent by participants and \prodscalpel on performing the three stages of SPL reengineering: feature \emph{extraction}, \emph{adaption} and \emph{merging}. The graph highlights the average time spent by participants who successfully generated new products.} 
	\label{fig:experiment_result_time_II}
\end{figure}

Most of the participants from Group B had not completed the feature migration process from \emph{Mytar}  within the 4 hours time limit. Considering the participants that were able to finish the process (i.e., participants \emph{P17}, \emph{P18} and \emph{P19}) successfully (all tests passed) they spent an average of 2h23 minutes while \prodscalpel was able to complete this task in 19 of 20 trials in the timeout, taking 27 minutes on average.

By considering the time spent in both scenarios, the tool accomplished the product line generation process 4.8 times faster than the mean time taken by participants who were able to finish the experiment within the timeout.

\textbf{Statistical analysis of performance.} 
To establish statistical significance of our results we first establish statistical distribution of runtimes for each scenario through the Shapiro-Wilk~\cite{SHAPIRO1965} test. 
Next, we use ANOVA~\cite{Gelman2005} and Pairwise Student’s t-test analysis to test the hypothesis that \prodscalpel's runtimes are statistically lower when compared with the times required by the participants to complete the same task (p-value $<$ 2e-16). We rejected the null hypothesis for all pairs. Figure~\ref{fig:experiment_result_boxplot} graphically shows the time results for our two groups in comparison with \prodscalpel's performance. In scenario I, the preliminary information provided by the box plots indicates that all samples are normally distributed (W = 0.70445, p-value = 3.129e-05).

In scenario II, the normality test result showed a normal distribution with a W = 0.69378, p-value = 1.715e-06. Thus, we used ANOVA to hypothesis testing and Pairwise t-Student. Based on the ANOVA test and Pairwise t-Student, we rejected the null hypothesis (p-value $<$ 2e-16) that the distribution of the population is homogeneous.

We can conclude that \prodscalpel reduces developer effort to transfer features to a product line in both scenarios. For both simulation scenarios, there is a significant effect size between the tasks performed in a manual way and using \prodscalpel. 
Our participants had similar performance times in scenario I, with the exception of participant \emph{P6}. On the other hand, most of the participants of scenario II do not terminate the experiment before the 4-hour timeout. This last one is qualitatively explained by the participants in the post-survey where they mentioned it is hard to adapt a feature from one codebase to run in a different codebase.

\begin{figure}[t]
	\centering \includegraphics[width=0.9\textwidth]{images/experiment_time_results.png}
	\centering 
	\caption{Time results grouping automated and manual in both scenarios.  Scenario I: \emph{NEATVI} - Product Base; Scenario II: \emph{Mytar} - Product base.}
	\label{fig:experiment_result_boxplot}
\end{figure}

\begin{framed}
\noindent \textbf{RQ4} The results confirmed that \prodscalpel outperformed participants transplanting two features 4.8 times faster, on average, compared to participants who finished the process within timeout.
\end{framed}

\section{Threats to Validity}

In this section we present threats to validity of our studies.

\textbf{External Validity.} The relatively small number and diversity of systems used for derive new products pose an external threat to validity. For our user study we applied our results to small programs due to the boundaries of an in-lab study; our results may not generalize to larger programs in the wild. We tried to mitigate it by constructing possible real-world scenarios, i.e., reuse of features from unrelated codebases and variations of the same real-world systems. Additionally, given that our approach was helpful even for small programs, we argue that it is likely helpful for larger systems as it is nearly impossible to incorporate new variants to a product line without a large understanding of the donor systems specifications~\cite{Assuncao2017} and without considerable adaption effort.

\textbf{Internal Validity.} Due to time expensive nature of a human study, we had only 20 participants. We tried to mitigate this issue by selecting participants with considerable experience in SPL projects. The other threat to the validity is the system size; small features are used in this experiment. We assume that inspecting the code to transfer a feature to a product line is hard, which has been confirmed in our post-study survey. We conducted a pilot study to mitigate threats arising from such issues.
Even with a limited execution time, we were able to transplant features from donors of significant size, considering the number of code lines of the subject systems, as shown in Table~\ref{tab:instrumentation}. 
Most participants were also able to complete the given tasks within the time limit.
We also use testing as means of validating our approach, which cannot provide a formal proof of correctness.  We used extensive testing to mitigate this risk. Moreover, testing is a standard approach to code evaluation in real-world scenarios due to its high scalability.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec08-related_work.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Related Work} \label{sec:related_work}

In this section, we present an overview of existing research related to our work. We position our work within the existing body of knowledge in areas of reengineering of systems into SPL, clone-and-own, variability in SPL and software transplantation.

\textbf{Reengineering of Software Systems into SPL}. Diverse academic proposals and industrial experience reports addressing reengineering of legacy systems into SPL are present in the literature~\cite{Assuncao2017}. However, this number decreases considerably when we are interested in proposals that automate the lifecycle of the reengineering process~\cite{Kruger2020}.

Martinez et al.~\cite{Martinez2015} introduced \emph{But4Reuse}, a generic and extensible open source tool to facilitate extractive SPL adoption. But4Reuse is a tool that aims to extract SPLs from legacy systems by identifying a set of reusable assets and representing them in a modular way. The tool uses a variety of program analysis techniques, including clone detection and feature location, to identify commonalities and variabilities in the code. Once the SPL is extracted, But4Reuse generates a set of variability models, which can be used to configure the SPL for different product variants. In contrast, \FOUNDRY does not assume an existing set of product variants. 
SPL can be created from a single codebase, and only requires feature entry point annotation, and a set of tests.
The needed code is automatically extracted using slicing, and modified to run in the given product base via automated over-organ adaptation.

\emph{IsiSPL}~\cite{hlad2021} is a reactive approach~\cite{Krueger2001} to SPL adoption. 
IsiSPL automates the integration of new products into an existing SPL and thus generation of a new SPL with the new features.
In particular, whenever a new product is added, a list of all features needs to be provided.
IsiSPL then analyses SPL to only insert new features, annotating them with conditional directives. However, with large number of products inserted over time, the list of conditional directives will grow, hindering code comprehension, maintenance, and ease of derivation of new products. In \FOUNDRY the use of such directives to handle variability is not a mandatory condition to use it. The developer is free to automatically introduce conditional directives in the evolved product line. That is they can also generate a product without surrounding the feature's code with additional directives, simply by transplanting organs from the transplantation platform.

\textbf{Clone-and-own}. \FOUNDRY can be exploited as an automated alternative to clone-and-own, where, instead of creating a product line, products are cloned and amended, based on demand.
Although there exists automated support for feature detection using code clone detection, its adaptation for reuse still requires manual work~\cite{Yoshimura2006, Kastner2014}.

Fischer et al.~\cite{Fischer2015}, for instance, present \emph{ECCO} to enhance clone-and-own. 
The tool finds the proper software artifacts to reuse and then provides guidance during manual adaptation phase, by hinting which software artifacts may need to be migrated and adapted. 
Moreover, ECCO requires that the features' source code must be extracted from the same family of products, which limits its ability to reuse assets. 
In contrast, \FOUNDRY stores over-organs that can be automatically adapted to different product bases. These do not need to come from the same family of products as the product base. Once extracted, features implemented by stored over-organs can be adapted, and implanted into a product base in a fully automated way.

\textbf{Variability in SPL.} The capacity of providing variability in a software development process is a key aspect of modern software development, enabling software products to be customized and adapted to meet the needs and requirements of different stakeholders.

Several works have already identified the frequent use of the \emph{variability mechanisms}, like preprocessor directives~\cite{Liebig2010} and feature flags~\cite{ Kruger2018, Meinicke2020}, as strategies for allowing the inclusion or exclusion of specific code blocks or features in the product line at compile time or runtime. Both are annotation-based implementation techniques for software product lines require explicit annotations often scattered across multiple code units (e.g., preprocessor annotations such as $\texttt{\#IFDEFs}$)~\cite{Apel2013}. 
These annotations establish a mapping of portions of code to features defined in a variability model. This mapping serves as input to the configurator tools, which then uses the information to select and configure the appropriate features for a given software product.

Despite its error-proneness and low abstraction level, the preprocessor directives are still widely used in present-day software projects to implement variability, maintain, evolve, reuse, or re-engineer a software system~\cite{Kruger2018}. Liegib et al.~\cite{Liebig2010} presents a study of 40 SPL that use preprocessor-based variability mechanisms. The study analyzes the variability mechanisms used in the product lines and the impact of these mechanisms on the codebase, in terms of code size, complexity, and maintainability. Christian Kästner et al.~\cite{Kastner2008} also discuss the concept of variability based on preprocessor directives in software product lines and how it affects the granularity of features. They present a case study of the Linux kernel to illustrate how the different levels of granularity in feature implementation can affect product line evolution and maintenance.

Other authors, such as, Jezequel et al.~\cite{Jezequel2022} present a case study of how feature models and feature toggles can be used in practice to manage variability in software systems. The authors describe how they used feature models to capture the commonalities and variabilities of a software product line and then translated them into feature toggles that could be used to enable or disable specific features at runtime.

Although useful, these traditional variability mechanisms have limitations~\cite{Bachmann2005, Liebig2010}. Preprocessor directives can lead to code bloat and reduced maintainability, while feature flags can add complexity and overhead to the codebase. Rahman et al.~\cite{Rahman2016} analyzed feature flag usage in the open-source code base behind Google Chrome, finding that feature flags are heavily used but often long-lived, resulting in additional maintenance and technical debt. Meinicke et al.~\cite{Meinicke2020} also discovered that despite the temporary nature of feature toggles and developers' initial intention to remove them, they tend to remain in the codebase unless compelled by policy or technical measures.

Particularity, building an SPL , where the number of options can grow considerably ~\cite{Lotufo2010,Xu2015}, the use of feature flags and/or preprocessor directives can lead to a large codebase in the emergent product line~\cite{Meinicke2020}. Foundry can be an interesting alternative to those traditional variability techniques. In contrast to the existing approaches where lots of annotations that are permanently added and their number increases over time, our solution requires only an annotation of the feature entry point and its insertion point in the product base. Thus, Foundry's approach has the potential to reduce code complexity and increased readability, as such any extra annotations are not required. Furthermore, Foundry keeps all reusable features of a product line (so-called organs) functional and physically separate, integrating them into the product base only when required for composing a new product.

Overall, ST technique can potentially offer several advantages over the use of traditional feature toggles, including simplified maintenance, reduced code complexity, and reduced risk of conflicts. Furthermore, it can be used in conjunction with existing variability mechanisms like preprocessor directives and feature flags, allowing developers to take advantage of the benefits of both approaches such as the possibility of enabling or disabling organs at runtime or compile-time.

\textbf{Software Transplantation}. As far as the literature on automated software transplantation is concerned, Petke et al.~\cite{Petke2014,Petke2018} were the first to transplant snipets of code from various versions of the same system to improve its performance, using genetic improvement~\cite{Petke18}. One year later, Barr et al. successfully transplanted a feature from one program into another~\cite{Barr2015}.

Stelios Sidiroglou-Douskos et al.~\cite{Sidiroglou2017} proposed another tool, CodeCarbonCopy (CCC), which can also transplant code automatically. CCC is a code-transferring tool from a donor into a host codebase. It implements a static analysis that identifies and removes irrelevant functionalists that are irrelevant to the host system. It has performed well in eight code transfers across six applications. However, the code redundancy problem still persists.  CCC is thus unable to handle multiple feature transplantation. Foundry on other hand, addresses this significant problem by exploiting code clone detection. Additionally, CCC inherits the limitations of its static analysis technique~\cite{Bessey2010} to identify the feature codes. It typically only looks at the code in isolation and does not consider the broader context in which the code is used, such as external dependencies or interactions with other features~\cite{Bessey2010, Ilyas2016}.

More recently, Liu et al.~\cite{Liu2018} introduced a method to transplant code from open source software. The validation results indicated that their method can substantially reduce the workload of programmers and is applicable to real-world open-source software. However their idea is also based on program slicing, it does not support the transplant of multiple organs to reengineering of systems into SPL. Furthermore, they still not provide any tool that support their method.

In contract to those works, we are the first to use automated software transplantation to automate SPL engineering tasks. Our approach and tool can transplant multiple organs to compose an SPL from existing donor systems. In the process, we have solved several issues, previously not considered in the software transplantation literature, such as code redundancy, organ dependence and feature interactions.

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec09-conclusion.tex
Upload the journal latex project
This is the version submitted. It is tagged as v3.0. The historic of changes can be accessed from the overleaf version.

\section{Conclusions and Future Work} \label{sec:conclusion}

In this work, we propose an approach, \FOUNDRY, and a tool, \prodscalpel, that use software transplantation to speed conversion to and maintenance of SPL.
Both approach and the tool have been validated through case studies. 
We generated two products through the transplantation of features extracted from three real-world systems into two different product bases. 
Moreover, we performed an experiment with SPL experts to compare our approach with manual effort.
We showed significant time effort improvements when using \prodscalpel. 
The tool accomplished product line generation process by migrating two features 4.8 times faster than the mean time spent by participants who were able to finish the experiment within the timeout.

We argue that the migration to SPL transplantation-based in contrast to a configuration-based software product line makes it easier to use in practice.
Our approach improves SPL maintainability by physically separating features from the product base.  
\FOUNDRY can be used both for \emph{extractive} or \emph{reactive} product line migration, as well as as a systematic Clone\&own strategy to specialize existing products.
\FOUNDRY avoids code duplication of feature implementations, while preserving feature behavior. 
It can automatically propagate feature changes. 
That is, it provides solutions for problems often cited in reengineering of systems into SPL literature.

Our evaluation studies provide initial evidence to support the claim SPLE 
using software transplantation, is a feasible and, indeed, promising direction for SPL research and practice.
However, more studies are needed to provide more evidence for generalisability of our approach, and to investigate its applicability in an industrial context.

### 2023-11-28 response/rebuttal.tex
Update rebuttal.tex with the response of reviewer point 2.3
Write the first draft to the comment 2.3 from Reviewer 2.

@@
 \begin{point}
-
-1) The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
+The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
 \end{point}
 \begin{reply}
-	We appreciate the reviewer's thoughtful consideration of our paper.
-We acknowledge the concern raised regarding the clarity of the problem definition and its alignment with the provided example.
-We understand the importance of a well-defined problem statement for a thorough evaluation of our proposed solution.
 
-The central problem we address in our paper revolves around the challenges inherent in Software Product Line (SPL) re-engineering.
-Specifically, we focus on the need for a systematic and automated approach to transplant reusable components (organs) from existing SPLs into a new product base.
-This process, which we term "software transplantation," aims to streamline SPL re-engineering by enhancing reusability and reducing redundancy.
-
-To provide a concrete illustration, we offer an example in the paper to elucidate the transplantation process.
-However, we acknowledge that there might be room for improvement in aligning this example more closely with the general problem definition.
-We revisited the example in our manuscript to ensure better coherence with the overarching problem statement.
-
-In Section 3, where we detail our software transplantation approach, Foundry, we will provide additional context to clarify the problem definition further.
-This clarification include a refined articulation of the challenges associated with SPL re-engineering that our approach aims to address.
-By doing so, we wait had enhanced the overall coherence between the problem definition and the proposed solution.
-
-We appreciate the reviewer's guidance on this matter, and we are committed to enhancing the clarity of our problem statement to facilitate a more accurate evaluation of our solution.
-
-Thank you for your valuable feedback, and we look forward to addressing any further concerns you may have.
 \end{reply}
@@
 \begin{point} 
-1) In the abstract, you say that this proposal is to re-engineer existing product into a SPL.
+    In the abstract, you say that this proposal is to re-engineer existing product into a SPL.
 However, the motivating example and the solution does not fit well with this.
@@
 \begin{reply}
-	TODO
+	We appreciate the reviewer's comments and have carefully considered the concerns raised.We can see how what we wrote was unclear; in particular, the motivation example needed elaboration, and the mention of GEdit, Evince, and NEdit is vague by does not proves information that they just individual products that share some common features but can be used to generate product lines from a re-engineering process.
+
+    In response, we have revised the motivating example to enhance clarity and coherence with the overarching problem statement.
+We have rewritten it to emphasize the challenges faced by open-source projects like GNOME and how our proposed approach, Foundry, addresses these challenges, providing a clearer and more focused illustration of our methodology.
 \end{reply}
+
+\begin{changes}
+
+    \textbf{MOTIVATING EXAMPLE}
+
+    Original:
+    \begin{itemize}
+        \item Mentioned GNOME as an open-source project.
+        \item Highlighted the challenge of reusing features due to separate developments.
+        \item Introduced mass customization and common platform principles of SPLE.
+    \end{itemize} 
+    Revised:
+    \begin{itemize}
+        \item Retained the essence of GNOME as an open-source project.
+        \item Emphasized the obstacle of manual efforts hindering seamless feature reuse.
+        \item Clarified the principles of SPLE for mass customization and a common platform.
+    \end{itemize}
+
+    PROBLEM DEFINITION CONTEXT
+
+    Original:
+    \begin{itemize}
+        \item Introduced the problem faced by the GNOME project in adopting SPL.
+        \item Outlined the scenario of augmenting GEdit with additional features.
+        \item Mentioned the selection of GEdit as the product base and identified target features.
+        \end{itemize} 
+    Revised:
+    \begin{itemize}
+        \item Maintained the essence of the problem in adopting SPL by GNOME.
+        \item Expanded on the scenario of text editor enhancement.
+        \item Emphasized the role of GEdit as the product base and detailed the target features.
+        \end{itemize}
+
+    PROBLEM DEFINITION
+
+    Original:
+    \begin{itemize}
+        \item Described the challenges of automated transplantation, including code extraction and adaptation.
+        \item Mentioned potential dependencies among transplanted organs.
+        \item Concluded with the importance of overcoming challenges for successful SPL-ST adoption.
+    \end{itemize} 
+    Revised:
+    \begin{itemize}
+        \item Retained the challenges of automated transplantation, emphasizing diverse donor codebases.
+        \item Emphasized the careful adaptation of features to the host system.
+        \item Highlighted potential dependencies among transplanted organs and the crucial role in SPL-ST adoption. 
+    \end{itemize} 
+\end{changes}
 

### 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec02-motivating_example.tex
Update section motivating_example.tex
Rewrite section: motivating example.

@@
 %\eb{from Leandro: I've rewritten this sentence by providing details of how GNOME could benefit from using SPLE principles as mass customisation and a common platform.}
+
+\textbf{ORIGINAL VERSION}
 
@@
 In the donor, engineers need to demarcate all feature entry points to transplant; a single annotation is sufficient for \prodscalpel to extract a feature.
-%\todo{remove: To prepare the host, GNOME engineers use \prodscalpel to extract a product base by removing all not desired features. }
 
@@
 GNOME engineers then run \prodscalpel on these inputs, once per feature, with
-%\ls{This listing is better.
-First, the important thing is to show what a user would \emph{actually} type!
-Doing so subtly shows that the system is real.
-This text is nothing real and, as such, it invalidates the sentence that proceeds it.
-Second, why define these switches?
-They're self-defining!
-What could the switch "donor\_folder" mean other that the path to the donor source code?
-I am leaving this todo here briefly for you to read it before I revert your change.
-I'm trying to avoid churn.
-I would like your help with which ones you think are mandatory.
-Many should be either defaulted or in a configuration file, not on the command line.
-Thinking more about that, maybe all are configuration, not CLI, and we should present a config file here with a minimal CLI invocation.} \eb{I'm with you Earl, all parameters of this command can be considered configuration.
-I can improve it in the next version of the ProducScalpel or during the journal review process.
-Do you prefer I remove this command for the moment?}
+
 \begin{quote}
@@
    \end{quote}
-%
-%\begin{lstlisting}[language=bash]
-%prodScalpel --seeds_file <transplantation_directory>/seed-1.in 
-%    --compiler_options <transplantation_directory>/CFLAGS
-%    --host_target <transplantation_directory>/productBase/insert_point_file.c 
-%    --donor_target <donor_directory>/Donor/organ_souce_file.c 
-%    --donor_folder <donor_directory>/Donor/
-%    --workspace <transplantation_directory>/ 
-%    --core_function <core_funtion_name> 
-%    --host_project <transplantation_directory>/ProductBase
-%\end{lstlisting}
-%
-%\ls{\lstinline+<root>+ is used without definition.
-How can the Donor and Host share the same root?
-That undercuts our contribution that the Donor can be unaware of the host!
-Some of these switches look like they can be defaults, like seeds and compiler\_options and NOT reported/shown here.}
 
@@
 Open-source projects, like GNOME, are an especially promising source of code for \FOUNDRY, so long as the donors and target hosts share compatible licenses.
+
+\textbf{Reviewer Point P 2.3:} \emph{Reviewer Point P 2.3 — 1) In the abstract, you say that this proposal is to re-engineer existing product into a SPL.
+However, the motivating example and the solution does not fit well with this.
+In Section 2 you say that the GNOME team could be interested in offering an SPL of text editors.
+ Then, you describe two other tools where you could take features from and ”transplant” it to a common code base generated from Gedit.
+This is confusing because when you are in an extractive SPL approach you have a set of products and you do not know from which one you should start building the code base for your SPL.
+Why Gedit and not Evince or Nedit?
+Are really those systems of a family of systems managed from the same organization or are they just individual products that share some common features}
+
+\textbf{REVISED}
+
+In the dynamic landscape of open-source development, projects such as GNOME \footnote{https://wiki.gnome.org/Projects} harbor an extensive collection of independently evolving programs.
+While these programs share common features, their separate development trajectories hinder seamless reuse across the entire portfolio for mass customization, necessitating manual efforts.
+Principles of Software Product Line Engineering (SPLE)\cite{Pohl2005}, embodying mass customization and a common platform, could empower the GNOME team to reuse a foundational technology base while tailoring products to individual customers.
+
+However, the adoption of SPLE by GNOME faces a significant obstacle.
+The intricate web of programs within GNOME, developed independently over time, lacks a common platform and a software development process grounded in mass customization.
+Consequently, providing customized products and effectively managing the commonality and variability of features becomes a great challenge.
+
+The GNOME project, inherently suitable for SPLE, is restrained by the substantial investment of time and resources required for the significant re-engineering needed to adopt SPL.
+This is where Foundry steps in as a transformative approach, poised to alleviate the re-engineering burden and make SPL adoption feasible for GNOME.
+
+To elucidate, consider a scenario where the GNOME team aims to swiftly generate a product line within the domain of text editors.
+This endeavor involves being able to enhance their current text editor, GEdit\footnote{https://wiki.gnome.org/Apps/Gedit}, with additional features.
+GEdit is selected as the product base, a shared substrate that, in the context of Foundry, serves as the host for transplanted features.
+
+In the process of selecting the product base, GEdit was chosen based on its widespread use and familiarity within the GNOME project.
+Recognizing the distinct functionalities of Evince and NEdit, the decision to focus on GEdit, a widely adopted text editor within the GNOME ecosystem, was made to establish a practical starting point for creating a product line.
+While Evince and NEdit were part of the considerations, the emphasis on GEdit aligns with the overarching objective of leveraging an existing, widely used tool to facilitate the adoption of SPL.
+It is important to note that these programs, namely GEdit, Evince, and NEdit, are individual products that share common features, rendering them potential contributors to the envisioned product line.
+
+Suppose the team targets three features—(1) side-panel, (2) split pane, and (3) presentation.
+To enrich GEdit with these features, two donors are identified: NEdit\footnote{https://sourceforge.net/projects/nedit/}, a multi-purpose text editor external to GNOME, and Evince \footnote{https://wiki.gnome.org/Apps/Evince}, a document viewer within GNOME but not an editor.
+
+Here's where \prodScalpel, the tool automates Foundry, takes center stage.
+Iteratively and incrementally re-engineering the codebase for SPL, \prodScalpel enables engineers to demarcate feature entry points in the donor, extracting features with a single annotation.
+The host preparation involves extracting a product base from an existing system by removing non-shared features, readying it for feature implantation.
+
+The engineers annotate the host to indicate implantation points for each target feature, referred to as "organs" in transplantation nomenclature.
+Executing \prodScalpel per feature involves a specific command-line operation.
+This command automatically extracts the specified feature's source code and its dependencies, referred to as "over-organs" using transplantation nomenclature.
+
+Figure 1 illustrates this transplantation process, showcasing a new text editor derived from features transplanted from different donors, with GEdit as the product base.
+Through feature models representing each donor system and product base evolution, \prodScalpel adeptly transplants features such as the side-panel from GEdit, split\_pane from NEdit, and presentation from GNOME's Evince renderer.
+
+In essence, Foundry, through \prodScalpel, opens avenues for large-scale feature reuse by seamlessly transplanting features from any program into a product line.
+For projects like GNOME, characterized by extensive open-source contributions, Foundry emerges as a promising solution, assuming compatibility in licenses between donors and target hosts.

### 2023-12-15 response/rebuttal.tex
write the first draft of the rebuttal document

--- 2023-11-28 response/rebuttal.tex
@@
-% LaTeX rebuttal letter example. 
-% 
-% Copyright 2019 Friedemann Zenke, zenkelab.org
-%
-% Based on examples by Dirk Eddelbuettel, Fran and others from 
-% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
-% 
-% Licensed under cc by-sa 3.0 with attribution required.
+\documentclass[a4paper,11pt]{letter}
 
-\documentclass[11pt]{article}
 \usepackage[utf8]{inputenc}
+
+\usepackage{hyperref}
+\usepackage{palatino}
+\usepackage{color}
+
 \usepackage{lipsum} % to generate some filler text
 \usepackage{fullpage}
+\usepackage{xcolor}
+\usepackage{xspace}
 
-% import Eq and Section references from the main manuscript where needed
-% \usepackage{xr}
-% \externaldocument{manuscript}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Comments %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{ifthen}
 
-% package needed for optional arguments
-\usepackage{xifthen}
-% define counters for reviewers and their points
+\newboolean{showcomments}
+\setboolean{showcomments}{true} % comment this line to deactivate comments
+\ifthenelse{\boolean{showcomments}}{
+  \newcommand{\nbc}[3]{
+    {\colorbox{#3}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}%
+    {\textcolor{#3}{\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}}}
+  \newcommand{\todo}[1]{\nbc{TODO}{#1}{blue}\xspace}
+}{
+  \newcommand{\nbc}[3]{}
+  \newcommand{\todo}[1]{}
+}
+\newcommand{\eb}[1]{\nbc{Earl}{#1}{orange}}
+\newcommand{\jp}[1]{\nbc{Justyna}{#1}{blue}}
+\newcommand{\ls}[1]{\nbc{Leandro}{#1}{blue}}
+\newcommand\re[1]{\nbc{To Review:}{#1}{green}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% End Comments %%%%%%%%%%%%%%%%%%%%%%%%%
+% Margins
+\topmargin=-0.75in
+\textheight=9.55in
+\oddsidemargin=0pt
+\textwidth=6.6in
+
+%\let\raggedleft\raggedright % Date on the left
 \newcounter{reviewer}
@@
 
+%%% Project Macros
+\newcommand{\FOUNDRY}{\textsc{Foundry}\xspace}
+\newcommand{\prodscalpel}{\textsc{prodScalpel}\xspace}
 % This refines the format of how the reviewer/point reference will appear.
 \renewcommand{\thepoint}{P\,\thereviewer.\arabic{point}} 
-
-% command declarations for reviewer points and our responses
-\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
-                  \section*{Reviewer \thereviewer}}
-
 \newenvironment{point}
@@
 
-\newenvironment{reply}
-   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
-   {\medskip \end{sf}}
-
-\newenvironment{changes}
-   {\medskip \noindent \begin{sf}\textbf{Changes\_made}:\  }
-   {\medskip \end{sf}}
-
 \newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
@@
 
+\newcommand{\bluetext}[1]{\begingroup\color{black}#1\endgroup}
+\newcommand{\response}[1]{\textbf{Response:} \begingroup\color{black}#1\endgroup}
+\newcommand{\perhaps}[1]{\begingroup\color{magenta}#1\endgroup}
+\newcommand{\changes}[1]{\textbf{Changes to the paper:} \begingroup\color{blue}#1\endgroup}
+
+\signature{\bluetext{\mbox{Leandro Souza}}}
+
 \begin{document}
 
-\section*{Response to the reviewers, TOSEM-2023-0120}
-% General intro text goes here
-We thank the reviewers for their critical assessment of our work. 
-In the following we address their concerns point by point.
+\begin{letter}{\textbf{Response to Reviewers' Comments}}
 
-% Let's start point-by-point with Reviewer 1
-\reviewersection
+  \opening{\bluetext{Dear Editor,}}
 
+  \bluetext{%
+    Please find enclosed our response to the reviewers of our paper entitled
+    ``Software Product Line Engineering via Software Transplantation TOSEM-2023-0120''.
+    We would like to thank you for the opportunity you gave us to revise and submit the new version of our manuscript.
+    We did our best to address the main issues, remarks, comments, and suggestions provided by the referees.
+    They are discussed in detail below, and highlighted in blue in the paper.
+    We would also like to gratefully acknowledge the reviewers for their efforts and their valuable feedback to the paper.
+    They highly contributed to improving the quality of the manuscript.
+
+  }
+  %\closing
+\bluetext{With our best regards,}
+
+  %\ps
+\bluetext{Leandro Souza}
+
+  \bluetext{(e-mail: \href{mailto:leandro.souza@ifba.edu.br}{leandro.souza@ifba.edu.br})}
+
+  \noindent\rule{\textwidth}{0.4pt}
+
+  \textbf{Reviewer \#1:}
+
+  \bluetext{%
+    We would like to gratefully acknowledge the reviewer for his/her comments, remarks, and suggestions on the submitted paper.
+    We did our best to address the main issues raised in the reviewer's comments.
+    They are discussed below.
+  }
+
+  [Insert Reviewer's positiver comments here]
+
+  \response{Thank you very much for this positive feedback.}
+
+%[Insert Reviewer's  comment here]
   % Point one description 
@@
 \end{point}
-\begin{reply}
-TODO
-\end{reply}
+
+\response{Thank you for your valuable comment.
+Regarding the order of transplantations illustrated in Figure 1, we acknowledge that ProdScalpel is designed to be agnostic to the order, meaning it can work effectively with any sequence.
+However, it's pertinent to note that different orders may influence the resulting substrate's compactness.
+The choice of order can impact the dependencies between transplanted features, potentially affecting the overall structure and size of the generated substrate.
+Analyzing the underlying dependency graph in various orders could indeed provide insights into optimizing the compactness of the substrate, and we appreciate your suggestion to explore this aspect further.
+
+We agree that investigating the impact of transplantation order on missed artifacts, especially concerning the underlying dependency graph, could be an interesting avenue for future research.
+This analysis could contribute to a more comprehensive understanding of how ProdScalpel operates in different scenarios and its implications for reusability in compilation processes.
+We will consider incorporating this aspect into our evaluation to enhance the robustness and applicability of ProdScalpel.
+Thank you for highlighting this potential avenue for improvement and analysis.}
+
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
 
-\begin{changes}
-    TODO
-\end{changes}
+\response{Thanks for your interesting contribution.
+Your interpretation aligns with the intended purpose of the organ-host wrapper in Foundry.
+The wrapper was indeed idealized to facilitate the integration and execution of transplanted features within the target host.
+It acts as an intermediary layer responsible for providing access to the organ from the target host, and its construction is automated based on a specified implantation point in the product base.
+This flexibility aims to accommodate diverse features from different sources and streamline their integration into the product line.
+
+As for the possibility of the product line becoming polyglot, the answer is affirmative.
+The organ-host wrapper can designed to be versatile., enabling the calling of new features as different programs or executable in various languages.
+This flexibility is a deliberate design choice, aiming to accommodate diverse features from different sources and facilitate their integration into the product line.
+
+Concerning the impacts on the host system, adapt the process to a polyglot product line, can introduce a level of indirection that may have performance implications, especially if external processes like "exec" are involved.
+Additionally, this would have direct impact the GP algorithm that must be redesigned to handle with different languages or execution environments.
+
+If the wrapper is responsible for integrating features with different languages or execution environments, there might be additional overhead associated with data conversion, interoperability, or adapting features to the host system's context.
+The overall runtime performance of the product line may be affected.
+Features invoked through the wrapper may experience a delay compared to direct invocations, and this delay can accumulate if numerous features are called sequentially.
+It would be advisable to conduct performance evaluations and optimizations to ensure that the polyglot nature of the product line does not unduly compromise its efficiency.
+Additionally, the specific impacts would depend on the nature and complexity of the transplanted features and the execution environment.
+
+To mitigate potential performance impacts, careful consideration of the wrapper's design and implementation is important.
+Performance optimizations, such as minimizing unnecessary computations, employing efficient conversion strategies, and caching frequently used features, are essential for enhancing the overall responsiveness of the polyglot product line.}
+
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
-\begin{changes}
-    TODO
-\end{changes}
+
+\response{Thank you for your insightful observation and question regarding Figure 8.
+In fact in the context of the donor preparation to the transplant, the choice between retain or remove a portion of code surrounded by directives in the donor depends on the relevance of the code to the target organ.
+If the portion of code are used as a substract to the organ correctly works, ProdScalpel retain the code in the donor removing only its directives.
+Otherwise, it remove both code and directives that delimite it.
+Thus, the process illustrated in Figure 8 results in a clean donor codebase, important for providing an organ free of directives for the transplantation process.
+
+It is important notes that, in the context of host preparation, the choice between keep or remove the features from the product base depends on the target product line.
+In this case, the reconfiguratior is used clean the host codebase aims to proide a product base free of features not relevants to the resulting product line.
+
+We appreciate your attention to this aspect of ProdScalpel's functionality and ensure that this distinction is now clearly articulated in the relevant sections of the manuscript.}
+
+\changes{In the context of the donor preparation for the transplant, the choice between retaining or removing a portion of code surrounded by directives in the donor depends on the relevance of the code to the target organ.
+If the portion of code is used as a substrate for the organ to work correctly, ProdScalpel retains the code in the donor, removing only its directives.
+Otherwise, it removes both the code and directives that delimit it.
+Thus, the process illustrated in Figure 8 results in a clean donor codebase, important for providing an organ free of directives for the transplantation process.
+
+It is important to note that, in the context of host preparation, the choice between keeping or removing features from the product base depends on the target product line.
+In this case, the reconfigurator is used to clean the host codebase, aiming to provide a product base free of features not relevant to the resulting product line.}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
-\begin{changes}
-    TODO
-\end{changes}
+\ls{Dear Justyna and Earl, I don't have the analysis required to answer his questions. I can analise the code and run the tests to answer he if you think important to do it.}
+
+\response{}
+
+\changes{}
 
@@
 
-\begin{reply}
-	TODO
-\end{reply}
+\ls{Justyna and Earl, in Section 5.3, we provide an overview of the case studies.
+However, to address the reviewer's suggestion for a more in-depth analysis and a taxonomy of features, we need to enhance the discussion regarding the characteristics of features that prodScalpel can handle and those with which it faces challenges.
+Achieving this is currently not possible with the data we have.
+To do so, I would need to attempt transplanting all features from VIM, categorizing them based on complexity, dependencies, or other relevant criteria.
+This would allow us to explicitly mention any products that couldn't be run due to missed instructions and discuss the reasons behind such cases, providing a more comprehensive understanding of prodScalpel's capabilities and limitations.
+For the moment, I can explicitly state the limitations posed by the technologies used in prodScalpel, such as Doxygen and TXL, based on features not productized, like SPELL\_CHECK.
+I will also discuss potential avenues for future improvement and research to overcome these limitations.
+Do you think that this would be enough to answer the comment or not?}
 
-\begin{changes}
-    TODO
-\end{changes}
+\response{}
 
-\subsection*{Minor}
+\changes{}
 
-\shortpoint{Section 2, Page 4, line 20 "a document viewer for multiple document formats that is part of GNOME, not not an editor." -> an additional "not" maybe it was a "but" in the first place.}
-\shortreply{ TODO.}
+\textbf{Minor review}
 
-\shortpoint{Figure 3 mentions two products P and Q but no products are visible on the figure.}
-\shortreply{ TODO.}
+\begin{point}Section 2, Page 4, line 20 "a document viewer for multiple document formats that is part of GNOME, not not an editor." -> an additional "not" maybe it was a "but" in the first place.
+\end{point}
+\response{}
 
-\shortpoint{I wonder whether Figures 1, 2, and 3 somehow may be redundant. Could they be merged? Is it a necessity to keep all these 3? (at least 1 and 2 separated?)}
-\shortreply{ TODO.}
+\changes{}
 
-% Begin a new reviewer section
-\reviewersection
+\begin{point}Figure 3 mentions two products P and Q but no products are visible on the figure.
+\end{point}
+\response{}
+
+\changes{}
+
+\begin{point}I wonder whether Figures 1, 2, and 3 somehow may be redundant. Could they be merged? Is it a necessity to keep all these 3? (at least 1 and 2 separated?)
+\end{point}
+\response{}
+
+\changes{}
+
+  \noindent\rule{\textwidth}{0.4pt}
+
+  \textbf{Reviewer \#2:}
+
+  \bluetext{%
+    We would like to gratefully acknowledge the reviewer for his/her comments, remarks, and suggestions on the submitted paper.
+    We did our best to address the main issues raised in the reviewer's comments.
+    They are discussed below.
+  }
+
+  [Insert Reviewer's positiver comments here]
+
+  \response{Thank you very much for this positive feedback.}
 
 \begin{point}
+The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
+\end{point}
 
-1) The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
-\end{point}
-\begin{reply}
-	We appreciate the reviewer's thoughtful consideration of our paper.
-We acknowledge the concern raised regarding the clarity of the problem definition and its alignment with the provided example.
-We understand the importance of a well-defined problem statement for a thorough evaluation of our proposed solution.
+\response{}
 
-The central problem we address in our paper revolves around the challenges inherent in Software Product Line (SPL) re-engineering.
-Specifically, we focus on the need for a systematic and automated approach to transplant reusable components (organs) from existing SPLs into a new product base.
-This process, which we term "software transplantation," aims to streamline SPL re-engineering by enhancing reusability and reducing redundancy.
-
-To provide a concrete illustration, we offer an example in the paper to elucidate the transplantation process.
-However, we acknowledge that there might be room for improvement in aligning this example more closely with the general problem definition.
-We revisited the example in our manuscript to ensure better coherence with the overarching problem statement.
-
-In Section 3, where we detail our software transplantation approach, Foundry, we will provide additional context to clarify the problem definition further.
-This clarification include a refined articulation of the challenges associated with SPL re-engineering that our approach aims to address.
-By doing so, we wait had enhanced the overall coherence between the problem definition and the proposed solution.
-
-We appreciate the reviewer's guidance on this matter, and we are committed to enhancing the clarity of our problem statement to facilitate a more accurate evaluation of our solution.
-
-Thank you for your valuable feedback, and we look forward to addressing any further concerns you may have.
-\end{reply}
-\begin{changes}
-    TODO
-\end{changes}
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-TODO
-\end{reply}
 
-\begin{changes}
-    TODO
-\end{changes}
+\response{}
+
+\changes{}
 
 \begin{point} 
-1) In the abstract, you say that this proposal is to re-engineer existing product into a SPL.
+    In the abstract, you say that this proposal is to re-engineer existing product into a SPL.
 However, the motivating example and the solution does not fit well with this.
@@
 This is confusing because when you are in an extractive SPL approach you have a set of products and you do not know from which one you should start building the code base for your SPL.
-Why Gedit and not Evince or Nedit?
-Are really those systems of a family of systems managed from the same organization or are they just individual products that share some common features.
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
+
+\response{We appreciate the reviewer's comment and would like to provide further clarification on the motivating example and the selection of Gedit as the base for our SPL generation proposal via software transplantation.
+
+Our intention with the motivating example was to illustrate an extractive SPL process, demonstrating that new product lines can be generated through the reengineering process by transplanting features from different systems into a common product base.
+Gedit was chosen as a base due to its availability and suitability for showcasing the proposed approach.
+The decision to use Gedit is arbitrary, and we acknowledge that other text editors like Evince and Nedit could be equally valid choices.
+
+Regarding the choice of \emph{GEdit} as the product base in our example, it is important to note that our proposal does not dictate a fixed starting point for the codebase.
+Instead, the choice of the initial product base (\emph{GEdit} in this case) is a pragmatic decision based on various factors, including the number of features required for the target product line already present in the host system candidate.
+The choice of \emph{GEdit} in the example is not prescriptive; other text editors like \emph{VIM}, \emph{VI} or \emph{NEdit} could have been chosen based on the project requirements.
+
+To address the reviewer's concern about the apparent lack of a family of systems managed by the same organization, we want to clarify that Gedit, Evince, and Nedit are indeed individual systems, not part of a cohesive family managed by a single organization.
+In the case of Evince, while it shares some common features with Gedit, the example suggests the transplantation of a feature not present in Gedit, highlighting the flexibility of the proposed approach.
+
+It's crucial to emphasize that the choice of Gedit as the product base was made for illustrative purposes, and any other text editor could be employed to generate a different product line.
+The decision might be influenced by factors such as the number of features required for the target product line that are already present in the host system candidate.
+The key takeaway is the demonstrated capability to construct SPLs by strategically selecting a product base and extracting features from diverse donor systems.
+
+We hope this clarification addresses the concerns raised by the reviewer and underscores the flexibility inherent in our SPL generation approach.}
+
+\changes{}
 
@@
 The example seems to be more on "copying" features from other products rather than re-engineering an existing SPL. 
-Also, you present in Fig.
-1 the donors and the  host using a feature model.
-Are they already SPLs according to that?
+\end{point}
+
+\response{We appreciate the reviewer's engagement with our proposal and the opportunity to clarify the essence of our approach.
+The perceived focus on 'copying' features might be a misinterpretation, and we would like to emphasize that our proposal involves a novel solution for reengineering existing systems into Software Product Lines (SPLs) through the innovative concept of software transplantation.
+
+In traditional SPL development, the emphasis is on creating a common codebase from scratch or adapting an existing system to meet the requirements of a product line.
+Our proposal takes a different route by introducing the concept of 'software transplantation,' a method that goes beyond mere copying of features.
+
+The transplantation process involves extracting features from existing systems (donors) and integrating them into a common product base, demonstrating a transformative approach to SPL reengineering.
+This method allows for the strategic selection of features from diverse sources, providing an automated way to construct product lines.
+
+Unlike conventional SPL methods that may involve extensive modification or rewriting of code, our approach minimizes the need for extensive redevelopment.
+It leverages existing functionality from various systems, enabling the creation of product lines with reduced effort and increased adaptability.
+
+Our proposal represents a departure from traditional SPL practices, offering a automated solution that amalgamates features from different systems into a cohesive and adaptable product line.
+We hope this clarification highlights the innovative nature of our approach and its distinction from a simple 'copying' of features.}
+
+\changes{}
+
+\begin{point} 
+Also, you present in Fig. 1 the donors and the  host using a feature model. Are they already SPLs according to that? 
+\end{point}
+
+\response{We appreciate the reviewer's observation and would like to clarify the role of the feature model in our presentation, particularly in Figure 1.
+The use of a feature model is intended for feature representation and organization rather than implying that the depicted systems are already SPLs.
+
+In our proposal, the feature model serves as a visual representation to help SPL engineer to manage and understand the features evailable to the software transplantation process.
+It provides a structured view of the features that can be extracted from donor systems and how they can be integrated into the common product base.
+
+The feature model does not presuppose that the depicted systems are inherently SPLs.
+Instead, it aids in illustrating the relationships and dependencies among features, facilitating the decision-making process during the transplantation.
+
+Our proposal focuses on the reengineering of existing systems into SPLs through the software transplantation concept.
+The feature model is a tool in this process, offering a convenient way to organize and visualize features.
+We hope this clarification addresses the concern, and we are open to further discussions or adjustments to enhance the clarity of our presentation.}
+
+\changes{}
+
+\begin{point} 
 Are you re-engineering an SPL from existing SPLs?
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
 
-\begin{changes}
-    TODO
-\end{changes}
+\response{We appreciate the reviewer's question and would like to clarify the objective and nature of our proposal, especially concerning the motivating example and the choice of systems for reengineering.
+
+Our approach, as presented in the motivating example with \FOUNDRY, is centered around the reengineering of individual systems into Software Product Lines (SPLs).
+The goal is to facilitate the extraction and transplantation of features from diverse sources, creating a unified product line that combines functionalities from different individual systems.
+
+We believe that this approach, rooted in the software transplantation paradigm, presents a novel solution for SPL generation by leveraging existing codebases, irrespective of whether they belong to a portifolio of systems managed by the same organization or are individual products that does not share common features.
+
+We hope this clarification addresses the concerns raised by the reviewer, and we remain open to further discussions to enhance the clarity of our proposal}
+
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
 
-\begin{changes}
-    TODO
-\end{changes}
+\response{The motivating example is more demonstrative and serves as a conceptual illustration of \FOUNDRY's potential.
+The motivating example featuring \emph{GEdit, Evince, and NEdit} is intended to provide a broad, illustrative view of \FOUNDRY's capabilities.
+It aims to show how the proposed approach can be applied to diverse systems, not necessarily the ones used in the formal evaluation.
+
+The text editors (\emph{Kilo, VI, CFlow, VI, and VIM}) mentioned in the introduction are part of a focused set chosen specifically for the formal evaluation.
+The formal evaluation with specific text editors is designed for a more focused and detailed examination of the transplantation viability and ProdScalpel's performance.
+These editors were selected based on certain criteria such as  complexity, size and number of features, to rigorously test and demonstrate the effectiveness of \FOUNDRY.}
+
+\changes{}
 
@@
 However, this should be cleared right from the beginning.
-Nevertheless, the main problem comes when the solution is described.
-Section 3 is not formally or rigorously described.
-There is no a single definition along the sections and only descriptive paragraphs are used.
-This is unusual in prestigious journals such as TOSEM.
-The way you describe your proposal is more for a workshop to present ideas rather to a solid result to be presented in a journal.
 \end{point}
 
-\begin{reply}
-	TODO
-\end{reply}
+\response{We appreciate the reviewer's feedback regarding the clarity of our solution, specifically the distinction between "Foundry" and "ProdScalpel." We acknowledge the importance of clearly articulating their roles from the outset.
 
-\begin{changes}
-    TODO
-\end{changes}
+Foundry is the software transplantation approach we propose for the reengineering of individual systems into Software Product Lines (SPLs).
+It represents the conceptual framework and methodology for systematically extracting, adapting, and combining features from diverse sources to create a product line.
+
+On the other hand, ProdScalpel is the practical realization of Foundry.
+It is the tool designed to implement and automate the steps outlined by Foundry.
+ProdScalpel serves as the operational instrument, facilitating the extraction of features from donor systems, their adaptation to fit the host system, and the subsequent generation of a product line.
+
+To provide a clearer narrative, we will enhance the introduction to explicitly introduce Foundry as the software transplantation approach and ProdScalpel as the tool that operationalizes this approach.
+This clarification aims to eliminate any confusion regarding their respective roles and establish a better understanding for readers.}
+
+\changes{In our proposed approach, Foundry is a conceptual approach for the reengineering of individual systems into SPL.
+Foundry represents a novel software transplantation approach, providing a systematic means of extracting, adapting, and combining features from diverse sources to construct a product line.
+This conceptual foundation allows for the generation of SPLs by leveraging features from existing codebases, fostering an automated reengineering process.
+
+ProdScalpel is a practical realization of Foundry for codebases written in C.
+.
+It is a tool implemented to automate the procedural steps delineated by Foundry.
+It can be used to make possible the extraction of features from donor systems, adapting them to integrate with the host system, and ultimately, automating the generation of product lines.
+As the practical counterpart to Foundry's conceptual underpinnings, ProdScalpel bridges the gap between theory and implementation, providing a tangible means for developers to engage with and execute the reengineering process in an automated manner.}
+
+\begin{point} 
+Section 3 is not formally or rigorously described. There is no a single definition along the sections and only descriptive paragraphs are used. 
+\end{point}
+
+\response{}
+
+\changes{}
 
@@
 
-\begin{reply}
-	TODO
-\end{reply}
+\response{}
 
-\begin{changes}
-    TODO
-\end{changes}
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
+
+\response{}
+
+\changes{}
 
@@
 \end{point}
-\begin{reply}
-	TODO
-\end{reply}
 
@@
 \end{point}
+\response{}
 
+\changes{}
+
+\end{letter}
 \end{document}

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/Reviews/TOSEM/rebuttal.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

\documentclass[a4paper,11pt]{letter}

\usepackage[utf8]{inputenc}

\usepackage{hyperref}
\usepackage{palatino}
\usepackage{color}

\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{soul}
\usepackage[normalem]{ulem}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Comments %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{ifthen}
\usepackage{amssymb}
\usepackage[switch]{lineno}%line number

\newboolean{showcomments}
\setboolean{showcomments}{true} % comment this line to deactivate comments
\ifthenelse{\boolean{showcomments}}{
  \newcommand{\nbc}[3]{
    {\colorbox{#3}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}%
    {\textcolor{#3}{\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}}}
  \newcommand{\todo}[1]{\nbc{TODO}{#1}{blue}\xspace}
}{
  \newcommand{\nbc}[3]{}
  \newcommand{\todo}[1]{}
}
\newcommand{\eb}[1]{\nbc{Earl}{#1}{orange}}
\newcommand{\jp}[1]{\nbc{Justyna}{#1}{blue}}
\newcommand{\ls}[1]{\nbc{Leandro}{#1}{blue}}
\newcommand\re[1]{\nbc{To Review:}{#1}{green}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% End Comments %%%%%%%%%%%%%%%%%%%%%%%%%
% Margins
\topmargin=-0.75in
\textheight=9.55in
\oddsidemargin=0pt
\textwidth=6.6in

%\let\raggedleft\raggedright % Date on the left
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip %\hrule
                  \section{\textbf{Response to Reviewer} \thereviewer}}
%%% Project Macros
\newcommand{\FOUNDRY}{\textsc{Foundry}\xspace}
\newcommand{\prodscalpel}{\textsc{prodScalpel}\xspace}
% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{P\,\thereviewer.\arabic{point}} 
\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent 
	{\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
	\ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
	\medskip \end{sf}}

\newcommand{\bluetext}[1]{\begingroup\color{black}#1\endgroup}
\newcommand{\response}[1]{\textbf{Response:} \begingroup\color{black}#1\endgroup}
\newcommand{\perhaps}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\changes}[1]{\textbf{Changes to the paper:} \begingroup\color{blue}#1\endgroup}

% Abbreviations
\newcommand{\ie}{\hbox{\emph{i.e.}}\xspace}

\signature{\bluetext{\mbox{Leandro Souza}}}

\begin{document}

\begin{letter}{\textbf{Response to Reviewers' Comments}}

  \opening{\bluetext{Dear Editor,}}

  \bluetext{%
    Please find enclosed our response to the reviewers of our paper entitled
    ``Software Product Line Engineering via Software Transplantation'', \#TOSEM-2023-0120.
    We would like to thank you for the opportunity you gave us to revise and submit the new version of our manuscript.
    We did our best to address the main issues, remarks, comments, and suggestions provided by the referees.
    They are discussed in detail below, and highlighted in \textcolor{blue}{blue} in the paper.
    We would also like to gratefully acknowledge the reviewers for their efforts and their valuable feedback to the paper.
    They highly contributed to improving the quality of the manuscript.

  }
  %\closing
\bluetext{With our best regards,}

  %\ps
\bluetext{Leandro Souza, Earl T. Barr, Justyna Petke, Eduardo S. Almeida and Paulo Anselmo M. S. Neto}

  \bluetext{(e-mail: \href{mailto:leandro.souza@ifba.edu.br}{leandro.souza@ifba.edu.br})}
  \bluetext{(e-mail: \href{mailto:leandro.souza@ifba.edu.br}{leandro.souza@ifba.edu.br})}

\newpage
% Begin a new reviewer section
\reviewersection

  \bluetext{%
    We would like to gratefully acknowledge the reviewer for his/her comments, remarks, and suggestions on the submitted paper.
    We did our best to address the main issues raised in the reviewer's comments.
    They are discussed below.
    Changes resulted from these comments are highlighted in \textcolor{blue}{blue} in the revised manuscript.
  }

%Reviewer Point P 1.1
\begin{point}
  The problem tackled with this work is challenging and one of the identified bottlenecks in the field of SPL.
This work proposes a quite complete solution that seems to perform well as shown by the evaluation.
I only have a few comments about this paper, which are more like questions I think than criticisms.
\end{point}

  \response{Thank you very much for this positive feedback.}

%[Insert Reviewer's  comment here]
  % Point one description

%Reviewer Point P 1.2 ----------------------------------------------------
\begin{point}
Figure 1 shows a sequence of transplatations, which I agree is easy to represent for such kind of succession, but is the order really important? In my opinion, probably not, except maybe when analysing the underlying dependency graph? Could it be interesting to analyse the impact of this analysis on missed artefacts when using ProdScalpel? (maybe to relate to reusability in compilation)
	\label{pt:foo}
\end{point}

\response{Thank you for your valuable comment. Regarding the order of transplantations illustrated in Figure 1, we acknowledge that ProdScalpel is designed to be agnostic to the order, meaning it can work effectively with any sequence. However, it's pertinent to note that the choice of order can impact the dependencies between transplanted features, potentially affecting the overall structure and size of the generated substrate. 
}

\changes{We have added a discussion about the order of transplantations in Section 2, at 8th paragraph.}

%Reviewer Point P 1.3 ----------------------------------------------------
\begin{point}
Section 3.2, Page 9, "To overcome these issues Foundry introduces an organ-host wrapper. This layer is responsible for providing access to the organ from the target host. It is automatically constructed on demand, according to a given implantation point in the product base."+
when I saw this, I wonder whether Foundry would only produce wrappers to call new features and thus, would it be possible that the product line becomes polygot (in the sense that the wrapper may only be a call to "exec" for instance whatever the program called)? 
It is probably not as simple as that, but could it be a viable solution? What would be the impacts on the host system (in term of performance for instance)?
	\label{pt:foo}
\end{point}

\response{Thanks for your interesting observation. Your interpretation aligns with the intended purpose of the organ-host wrapper in Foundry, which was indeed idealized to facilitate the integration and execution of transplanted features within the target host. It acts as an intermediary layer responsible for providing access to the organ from the target host, and its construction is automated based on a specified implantation point in the product base. This flexibility aims to accommodate diverse features from different sources and streamline their integration into the product line.

As for the possibility of the product line becoming polyglot, the answer is yes. 
The organ-host wrapper can be designed to be versatile, enabling the calling of new features as different programs or executable in various languages. This flexibility is a deliberate design choice, aiming to accommodate diverse features from different sources and facilitate their integration into the product line.
Concerning the impact on the host system, adapting the process to a polyglot product line can introduce a level of indirection that may have performance implications, especially if external processes like "exec" are involved. Additionally, this would have direct impact on the GP algorithm that must be redesigned to handle different languages or execution environments.

If the wrapper is responsible for integrating features with different languages or execution environments, there might be additional overhead associated with data conversion, interoperability, or adapting features to the host system's context. The overall performance of the product line can be affected. Features invoked through the wrapper may experience a delay compared to direct invocations, and this delay can accumulate if numerous features are called sequentially. It would be advisable to conduct performance evaluations and optimizations, such as minimizing unnecessary computations, employing efficient conversion strategies, and caching frequently used features, to ensure that the polyglot nature of the product line does not unduly compromise its efficiency. 
}

\changes{We have added this observation to Section 4.3, at 4th paragraph.}

%Reviewer Point P 1.4 ----------------------------------------------------
\begin{point}
Section 4.1, page 12, "Figure 8 gives a example of a portion of code after prodScalpel cleaned up unused directives."
in the end, from what I can see from Figure 8, ProdScalpel is able to provide two output "partial source code", one to use when the feature is activated and the second one when it is not, when adding a new feature to the host. The choice between the two will depend on the new product to derive I assume. Is that correct?
	\label{pt:foo}
\end{point}

\response{Thank you for your insightful observation and question regarding Figure 8. To be more accurate, in the context of the donor preparation for the transplant, the choice between retaining or removing a portion of code surrounded by directives in the donor depends on the relevance of the code to the target organ. If the portion of code is used as a substrate to the organ and functions correctly, ProdScalpel retains the code in the donor, removing only its directives. Otherwise, it removes both the code and the directives that delimit it. Thus, the process illustrated in Figure 8 results in a clean donor codebase, which is important for providing an organ free of directives for the transplantation process.
}

\changes{We have added the explanation in Section 4.1, at 2nd paragraph.}

%Reviewer Point P 1.5 ----------------------------------------------------
\begin{point}
Section 5.2, page 15, when authors use VI and VIM both as donor and product base. Did authors took two different versions of VI (and VIM)? One with the desired feature and one without it. That could make sense to the why trying to transplant a feature to the same codebase. It could also be a maner to evaluate the tool. Is it able to provide the same code in the end than the original? How does it differ? Which tests pass or fail? etc.
	\label{pt:foo}
\end{point}

\response{We appreciate the reviewer's comment. We used VI and VIM both as donors and product bases, we used two different versions for each: one with the desired feature, as donor, and one without it, as host.

In our evaluation, ProdScalpel successfully replicated the original code with the transplanted feature, albeit with the addition of the organ-host wrapper. Tests carried out on the resulting code confirmed that the transplanted organ passed all acceptance tests, validating the successful integration of the feature into the original codebase.}

\changes{We have clarified the experimental setup in Section 5.3, at 4th paragraph.}

%Reviewer Point P 1.6 ----------------------------------------------------

\begin{point}
Section 5.3, page 16, when answering RQ2, I have the feeling that it lacks some in-depth analysis. From my understanding, I assume that some products cannot be run due to missed instructions or something else, but it is not discussed. The text only refers to documentation that is imprecise.
In addition, maybe RQ2 is expressed in a misleading way when put in light of the analysis. I was expected to get a taxonomy of different features that can or cannot be handled yet.

	\label{pt:foo}
\end{point}

\response{Thank you for your constructive feedback. We have revised RQ2 to better reflect the analysis conducted.
}

\changes{We have revised RQ2 to state: ``How effective is \prodscalpel~at productizing software?''}

%Reviewer Point P 1.7 ----------------------------------------------------

\begin{point}Section 2, Page 4, line 20 "a document viewer for multiple document formats that is part of GNOME, not not an editor." -> an additional "not" maybe it was a "but" in the first place.
\end{point}
\response{Thank you for bringing this to our attention. You are correct; it was indeed an error.}

\changes{
The corrected sentence now reads: "a document viewer for multiple document formats that is part of GNOME, but not an editor."}

%Reviewer Point P 0.8 ----------------------------------------------------
\begin{point}Figure 3 mentions two products P and Q but no products are visible on the figure.
\end{point}
\response{Thank you for bringing this to our attention. We acknowledge the error in the figure reference and appreciate your diligence in reviewing the text. The description you cited indeed belongs to Figure 2, not Figure 3. This has been corrected in the new version of the paper.}

\changes{The figure description has been moved to Figure 2.}

%Reviewer Point P 0.9 ----------------------------------------------------
\begin{point}I wonder whether Figures 1, 2, and 3 somehow may be redundant. Could they be merged? Is it a necessity to keep all these 3? (at least 1 and 2 separated?)
\end{point}
\response{Thank you for your thoughtful consideration of Figures 1, 2, and 3. We believe that maintaining these three figures separately is important for providing a comprehensive understanding of our approach. Each figure serves a different purpose in elucidating different aspects of the approach.

Figure 1 is designed to emphasize the iterative and incremental nature of our approach, showcasing the dynamic and evolving characteristics of the process over time. This figure contributes to a clearer visualization of the overall progression.

In Figure 2, our goal is to illustrate how the concept of software transplantation is employed to introduce variabilities and commonalities into the product line. This figure plays a pivotal role in conveying the foundational idea behind our approach.

Lastly, Figure 3 is dedicated to detailing each stage of the domain engineering process as outlined in our approach. By breaking down the process into stages, this figure provides a more granular view, helping readers comprehend the sequence of tasks in our approach.

While we recognize that merging figures might be a consideration, we believe that keeping them separate enhances the clarity and depth of the exposition. The distinct focus of each figure contributes to a more comprehensive understanding of our approach.}

\newpage  
% Begin a new reviewer section
\reviewersection

  \bluetext{%
    We acknowledge the time and effort the reviewer spent reading and commenting on the submitted paper.
    We have done our best to address the main issues raised.
    They are discussed below.
    Changes resulted from these comments are highlighted in \textcolor{blue}{blue} in the revised manuscript.
  }

\begin{point}
 The paper is interesting in the sense that tries to find synergies between two separate areas (SC and SPLE) to improve the SPLE process. It is also interesting that the proposal has been implemented and evaluated. The writing is in general easy to follow and the paper is well structured. 
 \end{point}

\response{Thank you for the positive feedback.}

\begin{point}
The general proposal is not clear. There are important mismatches in the problem definition and the example. Without a clear problem definition, the solution is difficult to evaluate. 
\end{point}

\response{Apologies the description of our approach has not been fully clear to you. Please note we identified four problems in SPLE in the paragraph at line 36 in the Introduction, and specify how our approach addresses those in the paragraph at line 50. We have addressed your particular points of confusion as follows:
}

\begin{point} 
    In the abstract, you say that this proposal is to re-engineer existing product into a SPL. However, the motivating example and the solution does not fit well with this. In Section 2 you say that the GNOME team could be interested in offering an SPL of text editors. Then, you describe two other tools where you could take features from and "transplant"  it to a common code base generated from Gedit. This is confusing because when you are in an extractive SPL approach you have a set of products and you do not know from which one you should start building the code base for your SPL. 
\end{point}

\response{Firstly, we did not state in the abstract that we ``re-enginer existing product into a SPL'', but that we ``re-engineer a codebase to support SPL''. Secondly, we would like to provide further clarification on the motivating example and the selection of Gedit as the base for our SPL generation proposal via software transplantation.

It’s important to emphasize that the choice of Gedit as the product base was made for illustrative purposes, and any other text editor could be employed to generate a different product line. Our intention with the motivating example was to illustrate an
extractive SPL process of text editors, demonstrating that new product lines can be generated through the reengineering process by transplanting features from different systems into a common product base, GEdit.

Regarding the choice of GEdit as the product base in our example, it is important to note that our proposal does not dictate a fixed starting point for the product base. Instead, the choice of the initial product base is a pragmatic decision based on various factors such as the number of features required for the target product line that are already present in the host system candidate.

To address the reviewer’s concern about the apparent lack of a family of systems managed by the same organization, we want to clarify that Gedit, Evince, and Nedit are indeed individual systems, not part of a cohesive family managed by a single organization. In the case of Evince, while it shares some common features with Gedit, the example suggests the transplantation of a feature not present in Gedit, highlighting one of the main strengths of our approach -- we can transplant features from multiple codebases.
}

%\changes{\jp{We have further clarified that .. Leandro - please expand so that the above clarifications also end up in the manuscript, thank you.} 
\changes{We have further clarified that in Section 2, at 1st, 2nd and 3rd paragraphs.}}

\begin{point} 
The example seems to be more on "copying" features from other products rather than re-engineering an existing SPL. 
\end{point}

\response{We appreciate the reviewer's observation and would like to clarify that the perceived focus on 'copying' features might be a misinterpretation, and we would like to emphasize that our proposal involves a novel solution for re-engineering existing systems into Software Product Lines (SPLs) through the concept of software transplantation.

Our focus is on transitioning to SPL and then maintaining that SPL, not re-engineering an existing SPL, as we state in our abstract and in the 5th paragraph of the Introduction, at line 50.  
Of course, we "copy" and then adapt features into a product, as is standard. 
The challenge lies in identifying, minimally extracting, and adapting such features to create new products (see, for instance: Wesley K. G. Assunção, Roberto Erick Lopez-Herrejon, Lukas Linsbauer, Silvia R. Vergilio, and Alexander Egyed. 2017. Reengineering legacy 868 applications into software product lines: a systematic mapping. Empirical Software Engineering 22 (2017), 2972–3016.). 
Our core contribution is precisely that we allow extracting and using features from code 
that is not officially participating in the SPL, and can even be unaware of the SPL, like 
features extracted from open source projects.

In traditional SPL development, the emphasis is on creating a common codebase from scratch or adapting an existing system to meet the requirements of a product line. Our proposal takes a different route by introducing the concept of 'software transplantation,' a method that goes beyond mere copying of features.

The transplantation process involves extracting features from existing systems (donors) and integrating them into a common product base, demonstrating a transformative approach to SPL reengineering. This method allows for the strategic selection of features from diverse sources, providing an automated way to construct product lines.

Unlike conventional SPL methods that may involve extensive modification or rewriting of code, our approach minimizes the need for extensive redevelopment. It leverages existing functionality from various systems, enabling the creation of product lines with reduced effort and increased adaptability.

Our proposal represents a departure from traditional SPL practices, offering an automated solution that amalgamates features from different systems into a cohesive and adaptable product line. We hope this clarification highlights the innovative nature of our approach and its distinction from a simple 'copying' of features.}

\changes{We have clarified this point in Section 1 and 4.}

\begin{point} 
Also, you present in Fig. 1 the donors and the  host using a feature model. Are they already SPLs according to that? Are you re-engineering an SPL from existing SPLs? 
\end{point}

\response{We appreciate the reviewer's observation and would like to clarify the role of the feature model in our presentation, particularly in Figure 1. The use of a feature model is intended for feature representation and organization rather than implying that the depicted systems are already SPLs -- they are not.

In our proposal, the feature model serves as a visual representation to help SPL engineer to manage and understand the features available to the software transplantation process. It provides a structured view of the features that can be extracted from donor systems and how they can be integrated into the common product base.

The feature model does not presuppose that the depicted systems are inherently SPLs. Instead, it aids in illustrating the relationships and dependencies among features, facilitating the decision-making process during the transplantation.

Our proposal focuses on the reengineering of existing systems into SPLs through the software transplantation concept. The feature model is a tool in this process, offering a convenient way to organize and visualize features. We hope this clarification addresses the concern.}

\changes{We have now provided further clarification in Section 2, at 7th paragraph, as follows:

"Importantly, the donor systems in this context are not product lines; we are using feature models to map some of their features, making it easier to understand what has been extracted from the donors' codebase."}

%\response{We appreciate the reviewer's question and would like to clarify the objective and nature of our proposal, especially concerning the motivating example and the choice of systems for reengineering.

%Our approach, as presented in the motivating example with \FOUNDRY, is centered around the reengineering of individual systems into Software Product Lines (SPLs). The goal is to facilitate the extraction and transplantation of features from diverse sources, creating a unified product line that combines functionalities from different individual systems.

%We believe that this approach, rooted in the software transplantation paradigm, presents a novel solution for SPL generation by leveraging existing codebases, irrespective of whether they belong to a portifolio of systems managed by the same organization or are individual products that does not share common features.

%We hope this clarification addresses the concerns raised by the reviewer, and we remain open to further discussions to enhance the clarity of our proposal}

%\jp{..till here}

\begin{point} 
Also, it is confusing in the example that you talk about three editors (gedit, evince and nedit). However, in the introduction you say that for evaluating your proposal you used other text editors (kilo, vi and cflow on top of vi and vim). This is confusing and distracts the reader. 
\end{point}

\response{The motivating example is more demonstrative and serves as a conceptual illustration of \FOUNDRY's potential. The motivating example featuring \emph{GEdit, Evince, and NEdit} is intended to provide a broad, illustrative view of \FOUNDRY's capabilities. It aims to show how the proposed approach can be applied to diverse systems, not necessarily the ones used in the formal evaluation.

The text editors (\emph{Kilo, VI, CFlow, VI, and VIM}) mentioned in the introduction are part of a focused set chosen specifically for the formal evaluation. The formal evaluation with specific text editors is designed for a more focused and detailed examination of the transplantation viability and ProdScalpel's performance. These editors were selected based on certain criteria such as  complexity, size and number of features, to rigorously test and demonstrate the effectiveness of \FOUNDRY.}

\changes{We have further clarified the criteria used in our choices for evaluation in Section 5.2, at 1st to 7th paragraphs.}

\begin{point} 
2) The solution is weakly presented. The difference between "foundary" and "prodscalpel" is confusing. After some reads, one can understand that one if the conceptual and the other is the implementation. However, this should be cleared right from the beginning. 
\end{point}

\response{We appreciate the reviewer's feedback regarding the clarity of our solution, specifically the distinction between "Foundry" and "ProdScalpel." We acknowledge the importance of clearly articulating their roles from the outset.

Foundry is the software transplantation approach we propose for the reengineering of individual systems into Software Product Lines (SPLs). It represents the conceptual framework and methodology for systematically extracting, adapting, and combining features from diverse sources to create a product line.

On the other hand, ProdScalpel is the practical realization of Foundry. It is the tool designed to implement and automate the steps outlined by Foundry. ProdScalpel serves as the operational instrument, facilitating the extraction of features from donor systems, their adaptation to fit the host system, and the subsequent generation of a product line.

To provide a clearer narrative, we enhanced the introduction by explicitly introducing Foundry as the software transplantation approach and ProdScalpel as the tool that operationalizes this approach. This clarification aims to eliminate any confusion regarding their respective roles and establish a better understanding for readers.}

\changes{We have introduced the following sentence in Section 1, at 9th paragraph, to better clarify the idea of Foundry and ProdScalpel.:

``Foundry is the conceptual solution for the reengineering of individual systems into SPL, while ProdScalpel is one possible implementation designed to automate the steps outlined by Foundry. As the practical counterpart to Foundry's conceptual underpinnings, Prodscalpel bridges the gap between theory and implementation.''}

\begin{point} 
Section 3 is not formally or rigorously described. There is no a single definition along the sections and only descriptive paragraphs are used. (..)
The solution is described in a vague way. A more theoretical or at least rigorous approach should be followed to describe the solution. 
\end{point}

\response{Thank you for your feedback. The primary objective of our paper is to present and describe our approach, along with the automation details through ProdScalpel. We intentionally focused on providing a comprehensive understanding of the software transplantation process for SPL re-engineering, emphasizing practical insights.

We would like to draw your attention to the fact that the transplantation process itself has been formally defined in our previous work titled "Automated Software Transplantation" ( reference 4 from the paper), which we mention in Introduction, paragraph 4 and Section 4, paragraph 1. In this paper, we aimed to bridge the conceptual understanding from the transplantation paper with the practical implementation presented here. Our intention was to balance theoretical foundation with practical insights to make the approach both conceptually clear and practically applicable in the context to the SPL reengineering. }

\begin{point} 
Why in Figure 2 system 4 is selected? why P and Q are the "code base" and feature N and O are removed? P and Q are not features that are shared with other systems, are they?
\end{point}

\response{Thank you for your observation. In Figure 2, the representation of features P, Q, N, and O are integral to illustrating the variability in the product line. These features are extracted from 3 donors systems into the product line and then implanted into the products that require each one of them. Features N and O signify features that are removed during the transplantation process, as they are deemed non-essential or irrelevant to the target product line or the derived products. On the other hand, features P and Q are retained in the product base and shared among the products in the product line. As a result, the product line includes features B, F, H, J, P, and Q, enabling the generation of diverse products through the software transplantation process.

To enhance the clarity of this aspect of the figure, we have included a paragraph providing detailed explanations for each component, aiming to ensure a more comprehensive understanding for readers. We appreciate your feedback and believe that these additional clarifications will address your concern.}

\changes{We have added the following clarifications in figure description and detail it in the Section 3,  at 3rd and 4th paragraphs:

"Figure 2 illustrates the concept under discussion. In the first part of the figure, we identify three donor systems from which features are transplanted into a codebase. The representation of features B, F, H, and J serves to demonstrate the inherent variability in the product line. These features are extracted into the product line and subsequently implanted into specific products according to their requirements.

In the second part of the figure, we have a product base generated from a host system. The features N and O in the host system symbolize elements that need to be removed during the transplantation process due to their irrelevance to the target product line or the resulting products. Features P and Q exemplify the commonality within the product line, as they will be kept in the product base and shared among multiple products. As a result, we have a product line comprising features B, F, H, J, P, and Q, enabling the generation of diverse products through the software transplantation process."}

\begin{point} 
In addition, the implementation is described as a tool paper and adds no value to rigorously and clearly present your approach/tool. 
\end{point}

\response{Thank you for your feedback. While we appreciate your perspective, we respectfully disagree with the notion that our implementation is purely a tool paper and lacks value in presenting our approach rigorously and clearly. In Section 4, we have made a deliberate effort to provide detailed insights into the implementation details of our tool, ProdScalpel. This includes how each stage of the transplantation process is automated and the strategies employed to address various challenges. By presenting the tool implementation, we aim to bridge the gap between theory and practice, offering a tangible demonstration of the concepts detailed in Section 3.

Moreover, the implementation details serve as a practical guide for researchers and practitioners interested in adopting our approach. The tool paper aspect complements the theoretical foundation provided in Section 3, contributing to the reproducibility and applicability of our approach in real-world scenarios. We believe that a comprehensive understanding of both the conceptual framework and its practical realization is essential for the successful adoption and adaptation of our approach in the broader software engineering community.}

\begin{point} 
- The title is confusing. It seems that the paper is more about "feature extraction and implantation using software transplantation" and it is not much about SPLE
\end{point}

\response{Thank you for your feedback. We would like to provide some clarification on your perspective on the title. While the the paper may emphasize the aspect of "feature extraction and implantation using software transplantation," it is important to recognize that our primary focus is on Software Product Line Engineering (SPLE) using software transplantation.

The term "feature extraction and implantation" is a fundamental component of our SPLE approach, where FOUNDRY employs software transplantation to systematically extract, adapt, and combine features from diverse sources to construct a product line. In essence, our research explores the intersection of software transplantation and SPLE, proposing a novel approach to SPL generation through automated reengineering.

The title aims to convey the integration of these two areas, emphasizing the unique contribution of software transplantation to SPLE. We believe that our approach, as presented in the paper, provides valuable insights into how software transplantation can be applied to enhance SPLE processes, addressing challenges related to maintainability, feature separation, and code duplication.

We hope this clarification helps in understanding the overarching theme of our research and its significance in the context of SPLE using software transplantation.}

\begin{point} 
In the abstract, the ideas are not clear. More details should be given. What is your solution about? In what is based? Only high level descriptions are given.
\end{point}

\response{Thank you for your feedback. Based on your comments, we have revised the abstract to provide more clarity on our solution. The abstract now offers additional details regarding the nature of our solution, focusing on \FOUNDRY's role in automating SPL processes through software transplantation. We have elaborated on the core concept of software transplantation and how it forms the foundation for \FOUNDRY's structured and automated approach to feature extraction and migration in SPLs. We hope these enhancements address your concerns and offer a clearer understanding of the paper's contributions.}

\changes{ The abstract now reads:

``Software Product Lines (SPLs) offer substantial benefits for companies producing related products by improving time-to-market, enhancing software quality, and reducing costs.  However, implementing SPL practices involves lengthy re-engineering and ongoing maintenance challenges. Current SPL practices rely on a suite of tools, each tailored for specific reengineering phases, leading to coordination and integration complexities for developers.
To address these challenges, we introduce \FOUNDRY, an automated approach utilizing software transplantation to expedite SPL conversion and maintenance. \FOUNDRY's solution is rooted in the concept of software transplantation, a process involving the adaptation and incorporation of code segments from one system into another. Building upon this foundation, \FOUNDRY systematically applies software transplantation to SPL, providing a structured and automated approach. By understanding the intricacies of feature extraction and migration, \FOUNDRY enables the generation of SPLs with increased speed and efficiency, addressing the challenges associated with manual efforts.
We realise \FOUNDRY in \prodscalpel, a tool that transplants multiple organs (\ie, a set of interesting features) from donor systems into an emergent product line for codebases written in C. To evaluate its effectiveness, we conducted an experiment comparing feature transplantation using \prodscalpel with manual effort. Our tool successfully produced two valid product lines, automating feature extrastion and integration from three real-world systems. The achieved results reveal that \prodscalpel achieved an average of 4.8 times faster feature migration across codebases compared to a group of SPL experts. ''}

\newpage
\begin{point} 
Add a reference to support the first sentence of the 2nd paragraph of Section 1
\end{point}

\response{
Thank you for the suggestion. In response to your comment, we added three references to support the first sentence of the second paragraph in Section 1. 
Here is how each reference supports the first sentence:

Breivold (2008): The reference by Breivold, titled "Migrating Industrial Systems towards Software Product Lines: Experiences and Observations through Case Studies," provides insights into the challenges and experiences of adopting Software Product Lines in industrial systems. It contributes to the understanding that upfront investment is a significant aspect of SPL adoption.

Bastos (2017): The reference to Bastos' work on "Software Product Lines Adoption in Small Organizations" offers perspectives on SPL adoption in small organizations. It supports the notion that upfront investment is a common consideration, even in smaller organizational contexts, reinforcing the general applicability of the statement.

Daniel (2020): The reference by Daniel, titled "Economic Impact of Software Product Line Engineering Method - A Survey," provides insights into the economic impact of Software Product Line engineering methods through a survey. In the paper, Daniel emphasizes that although the adoption of an SPLE might have plenty of benefits for an organization, it also implies several challenges; among them is the need for an initial cost investment. This aligns with our statement about the upfront investment in SPL adoption.}

\changes{The revised sentence now reads:

``Despite its benefits, adopting SPL requires considerable upfront investment before its benefits can be realized (references 7, 11 and 16 from the paper).''}

\begin{point} 
The first time that prodscalpel is mentioned is in line 44 (pag 3 of 32) and it is confusing that it is. 
\end{point}

\response{Thank you for your feedback. To enhance clarity, we will make adjustments to the text, ensuring a smoother introduction of ProdScalpel earlier in the abstract. }

\changes{The abstract now contains the following sentence:

"By understanding the intricacies of feature extraction and migration, FOUNDRY enables the generation of SPLs with increased speed and efficiency, addressing the challenges associated with manual efforts. We realise FOUNDRY in PRODSCALPEL, a tool that transplants multiple organs (\ie, a set of interesting features) from donor systems into an emergent product line for codebases written in C."}

\begin{point} 
Fig 1 is in page 4 but the first mention to it is in pag 5 and in another section. This is confusing. 
\end{point}

\response{Thank you for bringing this to our attention. We acknowledge the confusion caused by the placement of Fig 1 in page 4 and its first mention on page 5 in a different section. We have addressed this concern by moving the image to the next page, ensuring that it aligns with its first mention.}

\changes{The figure has been moved to the next page.}

\begin{point} 
Foundary is described as "an approach" (l 29 p 4). It would be nice to more precisely describe it: process? methodology fragment? tool? what?
\end{point}

\response{Thank you for your feedback. Foundry is the conceptual solution for the reengineering of individual systems into Software Product Lines (SPLs), while ProdScalpel is just one possible implementation designed to automate the steps outlined by Foundry.}

\changes{We have introduced the following sentence in Section 1, at 9th paragraph, to better clarify the idea of Foundry and ProdScalpel.

"Foundry is the conceptual solution for the reengineering of individual systems into SPL, while ProdScalpel is one possible implementation designed to automate the steps outlined by Foundry. As the practical counterpart to Foundry's conceptual underpinnings, Prodscalpel bridges the gap between theory and implementation."}

\begin{point} 
The statement that GNOME is a natural candidate for SPL is quite strong! (l 9, p5)
\end{point}
\response{Thank you for bringing this point to our attention. We recognize that the term "natural candidate" may convey a level of certainty that could be perceived as overly strong. In light of your suggestion, we have revised the statement to express a more nuanced perspective.
We hope this adjustment aligns better with the intended meaning.}

\changes{We omit the statement and rewrite the 1st, 2nd, and 3rd paragraphs of the "Motivating Example" section.}

\end{letter}
\end{document}

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/Reviews/TOSEM/aeletter.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

\documentclass[a4paper,10pt]{letter}

\usepackage[margin=1in]{geometry}
\usepackage{color}
\newcommand{\jp}[1]{\textcolor{blue}{#1}}

\name{Leandro O. Souza, Earl T. Barr, Justyna Petke, Eduardo S. Almeida, Paulo Anselmo M. S. Neto} % authors

% Signature of sender
\signature{\vspace{-3em} Authors of \#TOSEM-2023-0120} % manuscript number

% Address of sender
\address
{
}

%-----------------------------------------------------------------------------%

\begin{document}

% Name and address of receiver
\begin{letter}
{
   Associate Editor handling \#TOSEM-2023-0120
}

% Opening statement
\opening{Dear Dr. Groce,}

% Letter body

We hope you are well.

We are writing to ask for advice on how to handle some of Reviewer 2 (R2)'s comments to ensure that we are, all of us --- you, the reviewers, and us --- are effectively spending our time on revisions.
In particular,

\begin{enumerate}

\item R2 states ``In addition, the implementation is described as a tool paper and adds no value to rigorously and clearly present your approach/tool.'' (Point 2.10).  We present both an empirical study and a tool, but, accepting the counterfactual that we were solely an empirical study, even in this case, the study relies heavily on the tool, so presenting its implementation shows the challenge of implementing our approach in practice and crucially facilitates its replication and reuse.

\item R2 fundamentally misunderstands what we do. Multiple times R2 states we ``re-engineer an existing SPL'' (Points 2.4, 2.5), which is not what we do.  We allow a new way to identify and compose SPL components, 
including components from foreign code bases.

\item R2 states that ``Section 3 is not formally or rigorously described. There is no a single definition along the sections and only descriptive paragraphs are used.'' (Point 2.8) Formal definitions are in our earlier work, which we clearly state in the paper. We note that R2 is not complaining that our informal descriptions are not self-contained.  As such, we feel that it would be misleading to formally define these notions that we borrow from previous work here, as doing so would imply --- falsely --- that they are or we believed them to be contributions of the current submission.

\end{enumerate}

In addition to seeking advice for how best to address these issues, we are also bringing your attention to them, as one tactic we are considering is to violate conventional practice and simply overtly disagree with R2 on these points.  In that case, R2 will recommend reject with high probability.

We are hoping that the fact that we've 
opened this dialogue with you about the issues will mean that, either you can recommend a conventional path forward where we can constructively address R2's comments, or, failing that and R2 does reject us again, 
you would be amenable to overruling R2 or seeking another reviewer.

% Closing statement
\closing{Thank you,}
\end{letter}
\end{document}

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/main.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

@@
 \usepackage{adjustbox}
+\usepackage{soul}
+
+\usepackage[normalem]{ulem}
 \input{abbreviations.tex}
@@
 \usepackage[switch]{lineno}%line number
-%\linenumbers
+\linenumbers
 \AtBeginDocument{%
@@
 \newcommand{\ls}[1]{\nbc{Leandro}{#1}{red}}
-\newcommand\re[1]{\nbc{To Review:}{#1}{green}}
+\newcommand\re[1]{\nbc{}{#1}{blue}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% End Comments %%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{abstract}
-For companies producing related products, a Software Product Line (SPL) is a software reuse method that improves time-to-market and software quality, achieving substantial cost reductions. 
+\eb{To be deleted after your review}
+\sout{For companies producing related products, a Software Product Line (SPL) is a software reuse method that improves time-to-market and software quality, achieving substantial cost reductions.
 These benefits do not come for free.
@@
 Moreover, we conducted an experiment comparing \FOUNDRY's feature migration with manual effort.
-We show that \FOUNDRY automatically migrated features across codebases 4.8 times faster, on average, than the average time a group of SPL experts took to accomplish the task.
+We show that \FOUNDRY automatically migrated features across codebases 4.8 times faster, on average, than the average time a group of SPL experts took to accomplish the task.}
+
+\textcolor{blue}{Software Product Lines (SPLs) offer substantial benefits for companies producing related products by improving time-to-market, enhancing software quality, and reducing costs.
+However, implementing SPL practices involves lengthy re-engineering and ongoing maintenance challenges.
+Current SPL practices rely on a suite of tools, each tailored for specific reengineering phases, leading to coordination and integration complexities for developers.
+To address these challenges, we introduce \FOUNDRY, an automated approach utilizing software transplantation to expedite SPL conversion and maintenance.
+\FOUNDRY's solution is rooted in the concept of software transplantation, a process involving the adaptation and incorporation of code segments from one system into another.
+Building upon this foundation, \FOUNDRY systematically applies software transplantation to SPL, providing a structured and automated approach.
+By understanding the intricacies of feature extraction and migration, \FOUNDRY enables the generation of SPLs with increased speed and efficiency, addressing the challenges associated with manual efforts.
+We realise \FOUNDRY in \prodscalpel, a tool that transplants multiple organs (\ie, a set of interesting features) from donor systems into an emergent product line for codebases written in C.
+To evaluate its effectiveness, we conducted an experiment comparing feature transplantation using \prodscalpel with manual effort.
+Our tool successfully produced two valid product lines, automating feature extrastion and integration from three real-world systems.
+The achieved results reveal that \prodscalpel achieved an average of 4.8 times faster feature migration across codebases compared to a group of SPL experts.}
+
 \end{abstract}

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec01-introduction.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

@@
 Software Product Line (SPL) is a systematic methodology for producing related software products from shared development assets~\cite{Pohl2005, Linden2007, Fischer2015}.
-%SPL organises a codebase into a core asset features that are shared across all products within the product line.
-Then, it defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
 SPL organises features as core assets that are shared across all products within the product line.
 Then, it defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
-%SPL organises a codebase into a \emph{product base}, a set of  features, then defines a \emph{dependency relation} over those features and a \emph{variability mechanism} that creates products from subsets of the features, subject to the dependency relation.
 By centralising features into a product base, SPL prevents cross-product feature drift, allowing all products to benefit from a feature's improvement, if they use it.
@@
 
-Despite its benefits, adopting SPL requires considerable upfront investment before its benefits can be realised. 
+Despite its benefits, adopting SPL requires considerable upfront investment before its benefits can be realised \cite{Breivold2008, Bastos2017, Daniel2020}.
 The cost of migrating existing products to SPL is lower than adopting SPL from scratch, making \emph{extractive}~\cite{Krueger2001} adoption more common, especially in companies with many software system variants in production~\cite{Berger2013}.
@@
 Automating these tasks remains an open challenge~\cite{Assuncao2017}.
-
-%\textcolor{magenta}{From Justyna to Leandro: I am missing here a paragraph that would summarise related work on automating reengineering tasks.}\ls{fixed}
 
@@
 \FOUNDRY is independent of the programming language, and supports SPL's \emph{domain engineering} and \emph{application engineering}~\cite{Clements2001} processes at the code level.
+
 It tackles each SPL reengineering task, easing some and automating others.
@@
 We realise \FOUNDRY in \prodscalpel, a tool that transplants multiple organs (\ie, a set of interesting features) from donor systems into an emergent product line for codebases written in C.
-%\ls{variability mechanism is used without definition, just below:  it needs to be defined with a citation.} 
+It can be used to make possible the extraction of features from donor systems, adapting them to integrate with the host product base, and ultimately, automating the generation of product lines.
 \prodscalpel also supports the use of existing variability mechanisms~\cite{Gacek2001} based on \emph{feature toggle}~\cite{Rahman2016} or \emph{preprocessor directives}~\cite{Kastner2008B}.
 It can surround implanted organs with feature flags, which permit enabling and disabling features, to facilitate its integration into an existing SPL codebase that uses them.
+
+\textcolor{blue}{\FOUNDRY is the conceptual solution for the reengineering of individual systems into SPL, while \prodScalpel is one possible implementation designed to automate the steps outlined by Foundry.
+As the practical counterpart to \FOUNDRY's conceptual underpinnings, \prodscalpel bridges the gap between theory and implementation.}
 

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec02-motivating_example.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

--- 2023-11-28 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec02-motivating_example.tex
@@
 
- \begin{figure*}[t]
-	\centering \includegraphics[width=\textwidth]{images/incremental_ST4.png}
-	\caption{Product derivation process using the \FOUNDRY approach.
-\textit{\prodscalpel transplants three features, in sequence, into the GEdit's product base to derive a new text editor after three iterations of organ transplantation.} }
-	\label{fig:incremental_pd}
-\end{figure*} 
  %COMMENT: \ls{The case for SPL needs to be elaborated:  we must argue that the various GNOME subprojects share sufficient code to define a product base.
@@
 
-The open-source GNOME project\footnote{https://wiki.gnome.org/Projects} encompasses a large portfolio of individual programs evolve as independently as possible from the rest.
+\textbf{ORIGINAL VERSION}
+
+%to delete
+\sout{The open-source GNOME project\footnote{https://wiki.gnome.org/Projects} encompasses a large portfolio of individual programs evolve as independently as possible from the rest.
 These programs share features, but because they are separately developed, their constituent features cannot be easily reused across its portfolio to provide mass customization, at least without much manual effort.
 The combination of mass customisation and a common platform, principles of Software Product Line Engineering (SPLE)~\cite{Pohl2005}, would allows to GNOME team reuses a common base of technology and, at the same time, to bring out products tailored individual customers.
-Without a common platform and a software development process base on mass customisation, it may be more difficult to the GNOME project provides customized products and effectively manage the commonality and variability of its features.
+Without a common platform and a software development process base on mass customisation, it may be more difficult to the GNOME project provides customized products and effectively manage the commonality and variability of its features.}
 
-The GNOME  project is a natural candidate for SPL, but the significant reengineering investment of time and resources have prevented it from adopting SPL.
-\FOUNDRY is transformative because it can be used to reduce this cost.
+%\textcolor{blue}{In the dynamic landscape of open-source development, projects such as GNOME \footnote{https://wiki.gnome.org/Projects} harbor an extensive collection of independently evolving programs.
+While these programs share common features, their separate development trajectories hinder seamless reuse across the entire portfolio for mass customization, necessitating manual efforts.
+Principles of Software Product Line Engineering (SPLE)\cite{Pohl2005}, embodying mass customization and a common platform, could empower the GNOME team to reuse a foundational technology base while tailoring products to individual customers.
+However, the adoption of SPLE by GNOME faces a significant obstacle.
+The intricate web of programs within GNOME, developed independently over time, lacks a common platform and a software development process grounded in mass customization.
+Consequently, providing customized products and effectively managing the commonality and variability of features becomes a great challenge.}
 
-We show how the GNOME team could use \prodscalpel to quickly generate a product line.
+\textcolor{blue}{In the dynamic landscape of open-source development, projects such as GNOME \footnote{https://wiki.gnome.org/Projects} harbor an extensive collection of independently evolving programs.
+While these programs share common features, their separate development trajectories hinder seamless reuse across the entire portfolio for mass customization, necessitating manual efforts.
+Principles of SPLE\cite{Pohl2005}, embodying mass customization and a common platform, could empower the GNOME team to reuse a foundational technology base while tailoring products to individual customers.
+However, the adoption of SPLE by GNOME faces a significant obstacle.
+The intricate web of programs within GNOME, developed independently over time, lacks a common platform and a software development process grounded in mass customization.
+Consequently, providing customized products and effectively managing the commonality and variability of features becomes a great challenge.}
+%to delete
+\sout{The GNOME  project is a natural candidate for SPL, but the significant reengineering investment of time and resources have prevented it from adopting SPL.
+\FOUNDRY is transformative because it can be used to reduce this cost.}
+
+\textcolor{blue}{However, the adoption of SPLE by GNOME faces a significant obstacle.
+The intricate portifolio of programs within GNOME, developed independently over time, lacks a common platform and a software development process grounded in mass customization.
+Consequently, providing customized products and effectively managing the commonality and variability of features becomes a great challenge.}
+
+%to delete
+\sout{We show how the GNOME team could use \prodscalpel to quickly generate a product line.
 Suppose project collaborators want to build a product line in the domain of text editors.
@@
 Assume that the GNOME team targets the following three features (1) $\texttt{side-panel}$, (2) $\texttt{split pane}$, and (3) $\texttt{presentation}$.
-They then identify two donors from which to transplant these features:  \emph{NEdit}\footnote{https://sourceforge.net/projects/nedit/}, a multi-purpose text editor that is not part of the GNOME portfolio, and \emph{Evince}\footnote{https://wiki.gnome.org/Apps/Evince}, a document viewer for multiple document formats that is part of GNOME, not not an editor.
+They then identify two donors from which to transplant these features:\emph{NEdit}\footnote{https://sourceforge.net/projects/nedit/}, a multi-purpose text editor that is not part of the GNOME portfolio, and \emph{Evince}\footnote{https://wiki.gnome.org/Apps/Evince}, a document viewer for multiple document formats that is part of GNOME, not not an editor.
+}
 
-\prodscalpel iteratively and incrementally re-engineers the codebase for SPL.
+\textcolor{blue}{To elucidate, consider a scenario where the GNOME team aims to swiftly generate a product line within the domain of text editors.
+This endeavor involves became their current text editor, \emph{GEdit}\footnote{https://wiki.gnome.org/Apps/Gedit}, in a product line with the possiblity to transplant additional features.
+GEdit is selected as the product base, a shared substrate that, in the context of Foundry, serves as the host for transplanted features.}
+
+\sout{Suppose the team targets three features—(1) $\texttt{side-panel}$, (2) $\texttt{split pane}$, and (3) $\texttt{presentation}$.
+To enrich GEdit product line with these features, two donors are identified: \emph{NEdit}\footnote{https://sourceforge.net/projects/nedit/}, a multi-purpose text editor external to GNOME, and \emph{Evince}\footnote{https://wiki.gnome.org/Apps/Evince}, a document viewer within GNOME but not an editor.}
+
+\textcolor{blue}{Here's where \FOUNDRY and \prodscalpel} can be used to automate the re-engineering process.
+\FOUNDRY iteratively and incrementally re-engineers the codebase for SPL.
 In the donor, engineers need to demarcate all feature entry points to transplant; a single annotation is sufficient for \prodscalpel to extract a feature.
-%\todo{remove: To prepare the host, GNOME engineers use \prodscalpel to extract a product base by removing all not desired features. }
 
-To prepare the host, GNOME engineers use \prodscalpel to extract a product base from an existing system by removing all features not shared across all products within the built product line
-
+To prepare the host, GNOME engineers use \prodscalpel to extract a product base from an existing system by removing all features not shared across all products within the built product line.
 Then, the engineers must annotate the host to indicate the implantation point for each target feature, or “organ” using transplantation nomenclature.
 GNOME engineers then run \prodscalpel on these inputs, once per feature, with
-%\ls{This listing is better.
-First, the important thing is to show what a user would \emph{actually} type!
-Doing so subtly shows that the system is real.
-This text is nothing real and, as such, it invalidates the sentence that proceeds it.
-Second, why define these switches?
-They're self-defining!
-What could the switch "donor\_folder" mean other that the path to the donor source code?
-I am leaving this todo here briefly for you to read it before I revert your change.
-I'm trying to avoid churn.
-I would like your help with which ones you think are mandatory.
-Many should be either defaulted or in a configuration file, not on the command line.
-Thinking more about that, maybe all are configuration, not CLI, and we should present a config file here with a minimal CLI invocation.} \eb{I'm with you Earl, all parameters of this command can be considered configuration.
-I can improve it in the next version of the ProducScalpel or during the journal review process.
-Do you prefer I remove this command for the moment?}
+
 \begin{quote}
@@
 
-\Cref{fig:incremental_pd} illustrates all transplantation iterations performed to generate a new product.
-It shows a new text editor derived from the transplant of features from different donors and using GEdit as a product base.
-Using feature models~\cite{Kang1990} to represent each donor system and the product base evolution, \prodscalpel first transplants the \emph{side-panel} feature, extracted from GEdit itself.
-This transplantation demonstrates that \prodscalpel can transplant features into a product base that comes from the same codebase.
-Next, \prodscalpel transplants the \emph{split\_pane} feature from NEdit.
-It shows how \prodscalpel manages to transplant features from distinct codebases, which is not possible without manual effort using the current state-of-art to SPL reengineering.
-Finally, \prodscalpel transplants the \emph{presentation} feature from GNOME's Evince renderer.
+ \begin{figure*}[t]
+	\centering \includegraphics[width=\textwidth]{images/incremental_ST4.png}
+	\caption{Product derivation process using the \FOUNDRY approach.
+\textit{\prodscalpel transplants three features, in sequence, into the GEdit's product base to derive a new text editor after three iterations of organ transplantation.} }
+	\label{fig:incremental_pd}
+\end{figure*}
+
+Figure 1 illustrates all transplantation iterations performed to generate a new product.
+It shows a new text editor derived from the transplant of features from different donors and using GEdit as a product base, using feature models to represent each donor system and the product base evolution.
+\textcolor{blue}{Importantly, the donor systems in this context are not product lines; we are using feature models to map some of their features, making it easier to understand what has been extracted from the donors' codebase.} \prodscalpel first transplants the \emph{side-panel} feature, extracted from GEdit itself.
+This transplantation showcases \prodscalpel's ability to transplant features into a product base from the same codebase.
+Next, \prodscalpel transplants the \emph{split\_pane} feature from NEdit, highlighting its ability to integrate features from diverse sources seamlessly into a unified product base.
+Finally, \prodscalpel transplants the \emph{presentation} feature from GNOME's Evince renderer, further highlighting its ability to integrate features from diverse sources seamlessly into a unified product base.
+
+\textcolor{blue}{It is imperative to emphasize ProdScalpel's agnostic nature towards the order of transplantations.
+ProdScalpel is designed to operate independently of the transplantation sequence, providing practitioners with the flexibility to tailor the process according to their specific requirements.
+However, it is noteworthy that while ProdScalpel accommodates any transplantation sequence, the choice of order may impact the resulting substrate's compactness by influencing interdependencies between transplanted features.
+}
 

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec03-approach.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

@@
 
-Conceptually, in \FOUNDRY, while the product base provides commonalities (i.e., common features) to the target product line, the variability (i.e., variant features) are provided by the transplantation process, as illustrated in \Cref{fig:product_for_transplantation}.
+Conceptually, in \FOUNDRY, while the product base provides commonalities (i.e., common features) to the target product line, the variability (i.e., variant features) are provided by the transplantation process.
 This idea opens new ways for SPLE area by automated construction of different products by transplanting multiple organs into a product base.
+
+\textcolor{blue}{\Cref{fig:product_for_transplantation} illustrates the concept under discussion.
+In the first part of the figure, we identify three donor systems from which features are transplanted into a codebase.
+The representation of features B, F, H, and J serves to demonstrate the inherent variability in the product line.
+These features are extracted into the product line and subsequently implanted into specific products according to their requirements.}
+
+\textcolor{blue}{In the second part of the figure, we have a product base generated from a host system.
+The features N and O in the host system symbolize elements that need to be removed during the transplantation process due to their irrelevance to the target product line or the resulting products.
+Features P and Q exemplify the commonality within the product line, as they will be kept in the product base and shared among multiple products.
+As a result, we have a product line comprising features B, F, H, J, P, and Q, enabling the generation of diverse products through the software transplantation process.}
 
@@
 	\centering 
-	\caption{An overview of how new products are derived from a product line based on ST.}
+	\caption{An overview of how new products are derived from a product line based on ST.\textit{Four over-organs (A, D, G, L) are extracted from three donor systems and kept in the transplantation platform, with product base consisting of 2 features shared across all products (P and Q).}}
+
 	\label{fig:product_for_transplantation}
@@
 	\centering  \includegraphics[width=\linewidth]{images/FOUNDRY_DOM6.png}
-	\centering \caption{Domain engineering process supported by FOUNDRY.
-\textit{Four over-organs (A, D, G, L) are extracted from three donor systems and kept in the transplantation platform, with product base consisting of 2 features shared across all products (P and Q).} }
+	\centering \caption{Domain engineering process supported by FOUNDRY.  }
 	\label{fig:foundry_dom}
@@
 After the postoperative stage, new iterations of organ transplantation can be performed; thus, in a stepwise and incremental way, a new product is derived as organs are transplanted into a product base.
+
+\FOUNDRY is an approach that combines a conceptual framework with a systematic methodology for the reengineering of individual systems into SPLs.
+It functions as a process aimed at extracting, adapting, and combining features from various sources to construct a product line.
+\FOUNDRY is implemented and operationalized through \prodscalpel.
+While Foundry serves as a conceptual foundation and methodology, \prodscalpel acts as the operational instrument, automating the steps outlined by Foundry.
+Together, they provide a structured and automated solution for the generation of SPL by transition of features from individual systems to a cohesive product line.

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec04-implementation.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

@@
 
+\textcolor{blue}{In the context of the donor preparation for the transplant, the choice between retaining or removing a portion of code surrounded by directives in the donor depends on the relevance of the code to the target organ.
+If the portion of code is used as a substrate for the organ to work correctly, ProdScalpel retains the code in the donor, removing only its directives.
+Otherwise, it removes both the code and the directives that delimit it.
+Thus, the process illustrated in \Cref{fig:reconfigurator} results in a clean donor codebase, important for providing an organ free of directives for the transplantation process.}
+
 \subsection{Automating Vein and Over-Organ Extraction}
-As optimisations, \prodscalpel extracts the vein from multiple files.
-\prodscalpel uses backward slicing to identity each function belonging to the vein and saves it in a copy of its souce file in the transplantation platform.
+As optimizations, \prodscalpel extracts the vein from multiple files.
+\prodscalpel uses backward slicing to identity each function belonging to the vein and saves it in a copy of its source file in the transplantation platform.
 Then, it includes all functions calling existing into vein source code in the \emph{array of over-organ statements} used by GP to achieve an organ from the over-organ~\cite{Barr2015}.
@@
 In the end, \prodscalpel synthesises a call to the extracted organ to execute and test it from the wrapper constructed.
+\textcolor{blue}{The organ-host wrapper can be designed to be versatile, enabling the calling of new features as different programs or executable in various languages.
+This flexibility is a deliberate design choice, aiming to accommodate diverse features from different sources and facilitate their integration into the product line.
+Adapting the process to such a polyglot product line can introduce a level of indirection that may have performance implications, especially if external processes like "exec" are involved. 
+Additionally, this would have direct impact on the GP algorithm that must be redesigned to handle different languages or execution environments.}
 

### 2024-04-09 paper/FOUNDRY-TOSEM_JOURNAL/sections/sec06-case_studies.tex
review rebuttal version
New version of the paper and rebuttal documents generated after Leandro and Justyna reviewes

@@
 \begin{quote}
-\textbf{RQ2.} \emph{Which features can \prodscalpel~productize?}
+\textbf{RQ2.} \emph{\textcolor{blue}{How effective is \prodscalpel~at productizing software?}}
 \end{quote}
-We discuss the characteristics of features that \prodscalpel~can and those it cannot yet handle.
+\textcolor{blue}{We conduct an initial empirical study and} discuss the characteristics of features that \prodscalpel~can and those it cannot yet handle.
 Given the inherent complexity of the process, it is important to clearly define its current limitations.
@@
 
-\textbf{Subject systems}. We chose two text editors, \emph{VI} and \emph{VIM}, as product bases.
+\sout{\textbf{Subject systems}. We chose two text editors, \emph{VI} and \emph{VIM}, as product bases.
 For donors we chose subjects from two different domains: code analysis software and text editors.
@@
 We reuse \emph{VI} and \emph{VIM} as donors, as well as another text editor, \emph{kilo}.
-This is to show that donors can come from the same system as the product base, but also different systems within the same application domain.
+This is to show that donors can come from the same system as the product base, but also different systems within the same application domain.}
 
-Our donors are presented in Table~\ref{tab:donors_list}.
-We identified the following features as possible desired features in a new editor: $\texttt{output}$ from CFLOW, $\texttt{enableRawMode}$ from kilo, $\texttt{vclear}$ from VI, and $\texttt{spell\_check}$ and $\texttt{search}$ from VIM.
-Table~\ref{tab:donors_list} presents more details on the subjects.
+\textcolor{blue}{\textbf{Subject Systems and Selection Criteria.}To define the context for performing the case study research, the first step was to identify the most suitable codebases for composing the product lines.
+The selection of subject systems for our case studies was based on four criteria aimed at providing a comprehensive evaluation of the proposed approach and tool.
+These selection criteria were based on the convenient sampling method Wohlin et al.~\cite{Wohlin2012}.}
 
-\textbf{Procedure and execution}. 
-Initially, we automatically remove dead-code from both donors and host codebases by using \prodscalpel. 
-We also reduced each host to its basic form removing all optional features. Thus, both donors and host are prepared for the transplantation process. 
-Given an organs' entry point for each organ in the donor systems provided by us, their target implantation points in the product base, and a set of test suites for each organ, \prodscalpel was used to localise and extract a set of organs from the donor, transform each organ to be compatible with the context of their target sites in the product base and implant the organs in the beneficiary's environment.
-Each automated organ transplantation process was repeated 20 times, due to the heuristic nature of the over-organ adaptation process.
-We computed the average number of lines of code transplanted and the average runtimes for the transplantation process.
+\textcolor{blue}{\textbf{(i) Diversity and variation:} The selection of subjects should aim to encompass a range of diversity and variation to provide a comprehensive understanding of the phenomenon under investigation.
+This can include diversity in terms of domains, sizes, complexity, functionality, or other relevant characteristics.}
 
-\textbf{Case study environment}. The runtimes for each transplantation were measured on an Intel Core 3.1 GHz Dual-Core Intel Core i5, with 16 GB memory running MacOS 10.15.4.
+\textcolor{blue}{\textbf{(ii) Representativeness:} The selected subjects should be representative of the broader population or the specific context under study.
+They should capture the characteristics and variations found in the target population or application domain.}
+
+\textcolor{blue}{\textbf{(iii) Feasibility and accessibility:} Practical considerations, such as availability of data, availability of use and modification, access to organizations or systems, and resources required for data collection, should be taken into account during subject selection.}
+
+\textcolor{blue}{\textbf{(iv) Established systems:} Established systems are typically used in real-world scenarios and have a significant user base.
+Studying such systems can provide insights into practical challenges and solutions.}
+
+\textcolor{blue}{Base on these criteria, we present the systems used in Table~\ref{tab:donors_list}.
+We chose two text editors, \emph{VI} and \emph{VIM}, as product bases.
+These text editors serve as the foundation upon which the transplantation process is performed.
+By selecting widely used and established text editors, we increase the relevance of your case study.}
 
@@
 \end{table}
+
+\textcolor{blue}{To illustrate that donors can come from different application domains, we included \textit{GNU Cflow} one of the donors.
+\textit{GNU Cflow}, being a call graph extractor from C source code, represents a distinct domain compared to the text editors.
+This demonstrates that the transplantation process can be applied to diverse types of software systems.
+In addition to using donors from different domains, we also included \textit{VI} and \textit{VIM} as donors as well as another text editor, \emph{kilo}.
+This highlights the possibility of using donors from the same system as the product base or from different systems within the same application domain.
+It showcases the flexibility of the transplantation process and its ability to handle variations within a specific domain.}
+
+We identified the following features as possible desired features in a new editor: $\texttt{output}$ from CFLOW, $\texttt{enableRawMode}$ from kilo, $\texttt{vclear}$ from VI, and $\texttt{spell\_check}$ and $\texttt{search}$ from VIM.
+Table~\ref{tab:donors_list} presents more details on the subjects.
+
+\textbf{Procedure and execution}. 
+Initially, we automatically remove dead-code from both donors and host codebases by using \prodscalpel. 
+We also reduced each host to its basic form removing all optional features. Thus, both donors and host are prepared for the transplantation process. 
+Given an organs' entry point for each organ in the donor systems provided by us, their target implantation points in the product base, and a set of test suites for each organ, \prodscalpel was used to localise and extract a set of organs from the donor, transform each organ to be compatible with the context of their target sites in the product base and implant the organs in the beneficiary's environment.
+Each automated organ transplantation process was repeated 20 times, due to the heuristic nature of the over-organ adaptation process.
+We computed the average number of lines of code transplanted and the average runtimes for the transplantation process.
+
+\textbf{Case study environment}. The runtimes for each transplantation were measured on an Intel Core 3.1 GHz Dual-Core Intel Core i5, with 16 GB memory running MacOS 10.15.4.
 
@@
 
+\textcolor{blue}{In the transplantation of features from two different versions of the same system - one with the desired feature as the donor and another without it as the host - \prodscalpel replicated the original code, incorporating the transplanted feature.
+The introduction of the organ-host wrapper, a necessary addition, did not compromise the overall integrity or functionality of the codebase.
+To thoroughly validate this integration, acceptance tests were conducted on the resulting code.
+These tests affirmed the successful assimilation of the transplanted organ into the original codebase.}
+
+\textcolor{blue}{The capability to selectively remove or reimplant features in a software product line, facilitated by tools like \prodScalpel, can provide adaptability to the product line, allowing dynamic adjustments to meet changing requirements or market demands, providing flexibility in tailoring product configurations without extensive manual efforts.
+In terms of maintenance and evolution, the process of reimplanting features can automate updates or modifications, replacing intricate manual interventions.
+The systematic reuse of features across different product variants reduces redundancy and promotes consistency.
+This removal and reimplantation functionality contributes to product line customization, facilitating tailored solutions for specific customer needs.
+Additionally, the automated process significantly reduces development time by minimizing manual efforts, ensuring a more efficient product line engineering process.
+Furthermore, it enhances consistency and cohesion by uniformly applying changes to features across product variants, mitigating the risk of inconsistencies or errors.}
+
 In summary, these case studies show initial evidence that software transplantation can be successfully used for building product variants automatically by combining features transplanted from real-world systems.
@@
 However, we tried to mitigate it by constructing a possible real-world scenario, i.e., transplantation of features that would be useful in production of new text editors.
+By selecting different subject systems allowed us to assess the effectiveness, applicability, and versatility of \prodscalpel in achieving products.
+We used subjects from different domains and with a wide range of sizes to give evidence that \prodscalpel can also be used to achieve a product line from a different set of usage scenarios.
 
